### 2021-08-25 Presentation_inr_ESS9.tex
Initial Upload

\documentclass[aspectratio=169]{beamer}
\usetheme{default}
\useinnertheme{circles}
\setbeamertemplate{navigation symbols}{}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\usepackage[style=authoryear, doi=false]{biblatex}  
\addbibresource{~/Zotero/Bibliothek.bib}
\renewcommand*{\bibfont}{\tiny}

\author{Malte Grönemann}
\title{Respondent- and Interview-level predictors of Item Nonresponse in the ESS 9}
\date{20 May 2021}

\begin{document}

\begin{frame}
\maketitle
\end{frame}

\begin{frame}{Introduction}
 Missing Data are Bad
 \begin{itemize}
  \item decreases effective sample size
  \item nonresponse bias
  \item indicator of overall data quality (satisficing) \footcite{krosnickResponseStrategiesCoping1991}
 \end{itemize}

 \bigskip

 A Typology of Missingness \footcite{deleeuwPreventionTreatmentItem2003}
 \begin{itemize}
  \item missing completely at random
  \item missing at random
  \item not missing at random
  \item missing by design
 \end{itemize}
\end{frame}

\begin{frame}{Item Nonresponse}
Item nonresponse: non-substantive answer or refusal by the respondent

\bigskip

 Question-Answer-Process\footcite{deleeuwPreventionTreatmentItem2003}: understanding $\rightarrow$ recalling information $\rightarrow$  eventually judging $\rightarrow$  eventually editing $\rightarrow$  eventually formating $\rightarrow$  communicating

 \bigskip

 Two types of item nonresponse
 \begin{itemize}
  \item Don't know
  \begin{itemize}
  \item not \textit{able} to answer: mental effort is too high
   \item strategy of satisficing
   \item likely to be MAR
  \end{itemize}
\item Refusal
\begin{itemize}
 \item not \textit{wanting} to answer: privacy and sensitivity
 \item likely to be NMAR
\end{itemize}
\item But: mechanisms not perfectly separate \footcite{shoemakerItemNonresponseDistinguishing2002}
\end{itemize}
\end{frame}

\begin{frame}{Predictors of Item Nonresponse I}
 Country
 \begin{itemize}
  \item considerable country differences
  \item no good explanations to date \footcite{meitingerPowerCultureItem2020}
  \item some composition effects \footcite{kochItemNonresponseEuropean2009}
 \end{itemize}

  Survey
 \begin{itemize}
  \item compositional differences between modes \footcite{messerDeterminantsItemNonresponse2012}
  \item design possibilities and likelihood of unintended skipping
 \end{itemize}
 \end{frame}

 \begin{frame}{Predictors of Item Nonresponse II}
  Interviewer
 \begin{itemize}
  \item considerable interviewer effects
  \item barely any explanations \footcite{pickeryImpactRespondentInterviewer1998}
  \item poor interviewer behaviour?: skipping questions or shortening questionnaire
 \end{itemize}

 Respondent
 \begin{itemize}
  \item ability: age, health, education
  \item women have higher nonresponse
  \item ethnic minorities, poor language skills, low literacy

  $\rightarrow$ reflection of broader patterns of inequality \footcite{meitingerPowerCultureItem2020}
  \item motivation: political interest, feeling of powerlessness
 \end{itemize}
\end{frame}

 \begin{frame}{Predictors of Item Nonresponse III}
 Interview
 \begin{itemize}
  \item sometimes gender matching \footcite{vercruyssenEffectSociodemographicMis2017}
  \item similar age vs older interviewer
  \item other people present during the interview \footcite{kupekDeterminantsItemNonresponse1998}
  \item perceived cooperation of respondent \footcite{tuSocialDistanceRespondent2007}
 \end{itemize}

 Questionnaire
 \begin{itemize}
  \item explicitly offering DK
  \item complexity
 \end{itemize}

 Question
 \begin{itemize}
  \item sensitivity: e.g. income and sexual behaviour
 \end{itemize}

\end{frame}

\begin{frame}{Aim}
\begin{itemize}
 \item item nonresponse is still not well understood
 \item most established predictors are respondent characteristics $\rightarrow$ not changeable
 \item the aim is of course to ensure and improve data quality in the future
 \item directing attention to controlable predictors
 \item fundamental problem: data availability
 \item $\rightarrow$ explore reasonable available predictors especially on the interview level
\end{itemize}

\end{frame}

\begin{frame}{Data and Methods}
 \begin{itemize}
 \item ESS Round 9, all countries
  \item number of refusals and DK $\rightarrow$ overdispersed count data
  \item complete questionnaire excluding questions affected by routing to ensure comparability
  \item nested structure: interviews by interviewers within countries
  \item $\rightarrow$ negative binomial regression with interviewer fixed effects, standard errors clustered by interviewer
 \end{itemize}

\end{frame}

\begin{frame}{Preliminary Results}
\includegraphics[scale=.35]{results.pdf}
\end{frame}

\begin{frame}{Issues}
 \begin{itemize}
  \item missing data of included variables are directly linked to value of DV $\rightarrow$ imputation?
  \item model comparisons?
  \item descriptives? (word limit...)
 \end{itemize}

\end{frame}

\begin{frame}{References}
\printbibliography
\end{frame}
\end{document}

### 2021-08-25 inr_ESS9.tex
Initial Upload

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage[a4paper,left=2.5cm,right=2.cm,top=2.5cm,bottom=2.5cm]{geometry}

\usepackage[english]{babel}
\usepackage[babel]{csquotes}

\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[all]{hypcap}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs}

\usepackage[style=authoryear, backend=biber, natbib]{biblatex}  
\bibliography{~/Zotero/Bibliothek.bib}

%opening
\title{Interview- and Respondent-level Predictors of Refusal and ``Don't know'' Item Nonresponse in the ESS 9}

\author{Malte Grönemann}

\begin{document}

\begin{titlepage}
\maketitle
\thispagestyle{empty}

%\centering
\large

Term Paper within the Seminar \\
\textbf{Fundamentals of Survey Design}  \\
FSS 2021 \\
Tobias Gummer\\
University of Mannheim \\
School of Social Sciences

\bigskip

\normalsize

\textbf{Abstract:} In this article, I have a look at interview- and respondent-level predictors of item nonresponse in the ESS Round 9. Using an interviewer fixed-effects negative binomial regression, I confirm most previous findings on the respondent-level: better educated, politically more interested and engaged, more trusting male ethnic majority respondents have lowest rates of item nonresponse. But the interview level is the more interesting one since these predictors actually can help survey researchers to prevent item nonresponse. Besides efforts to ensure respondents understand the survey questions, building a trusting interview situation to ease reluctancy, stimulating motivation and making sure that respondents correctly use showcards are the most promising avenues to reduce item nonresponse. Gender matching and interferences with the interview are not associated with item nonresponse.

\bigskip

\textbf{Keywords:} Item Nonresponse, Data Quality, European Social Survey, F2F, Interview, Fixed-Effects Negative Binomial Regression

\end{titlepage}

\section*{Declaration in Lieu of an Oath}

\setcounter{page}{0}
\thispagestyle{empty}

„Hiermit versichere ich, dass diese Arbeit von mir persönlich verfasst ist
und dass ich keinerlei fremde Hilfe in Anspruch genommen habe. Ebenso
versichere ich, dass diese Arbeit oder Teile daraus weder von mir selbst
noch von anderen als Leistungsnachweise andernorts eingereicht wurden.
Wörtliche oder sinngemäße Übernahmen aus anderen Schriften und
Veröffentlichungen in gedruckter oder elektronischer Form sind
gekennzeichnet. Sämtliche Sekundärliteratur und sonstige Quellen sind
nachgewiesen und in der Bibliographie aufgeführt. Das Gleiche gilt für
graphische Darstellungen und Bilder sowie für alle Internet-Quellen. Ich
bin ferner damit einverstanden, dass meine Arbeit zum Zwecke eines
Plagiatsabgleichs in elektronischer Form anonymisiert versendet und
gespeichert werden kann. Mir ist bekannt, dass von der Korrektur der
Arbeit abgesehen und die Prüfungsleistung mit „nicht ausreichend“
bewertet werden kann, wenn die Erklärung nicht erteilt wird."
\bigskip

"I hereby declare that the final thesis presented is my own work and that I have not called upon the help of a third party. In addition, I affirm that neither I nor anybody else has submitted this paper or parts of it to obtain credits elsewhere before. I have clearly marked and acknowledged all quotations or references that have been taken from the works of others. All secondary literature  and  other  sources  are  marked  and  listed  in  the  bibliography.  The  same  applies  to  all charts, diagrams and illustrations as well as to all Internet resources. Moreover, I consent to my paper being electronically stored and sent anonymously in order to be checked for plagiarism. I am aware that if this declaration is not made, the thesis may not be graded."

\bigskip

Mannheim, \today

\bigskip
\includegraphics[scale=.1]{/home/malte/Dokumente/Dokumente/Unterschrift_Malte.JPG}
\newpage

\section{Introduction}

Missing data pose a problem to analysis of survey data because they decrease the effective sample size and can introduce bias to estimations if the causes for missingness are related to the missing item or respondent characteristics \citep{deleeuwPreventionTreatmentItem2003}. Due to these harmful effects of missing data, one objective of survey researchers is to keep their prevalence as low as possible. For this aim, it is important to understand the processes that can lead to missing data.

We can get a first grasp on both the mechanisms causing the missingness and its likely impact on bias by categorising missingness based on their relation to randomness. Missings that are not related to anything are typically called \textit{missing completely at random} and have no impact on bias therefore only decreasing effective sample size. Is the relation between missingness and item not direct but the cause of missingness is related to other characteristics of the respondent, we speak of \textit{missing at random}. MAR additionally has the risk of leading to biased estimates. Even more consequential for bias are data \textit{not missing at random} where the missingness is directly linked to the unknown position to the surveyed topic or unknown value of the item. There is a special case of not missing at random in missings though that is unconsequential because it is controlled: \textit{missing by design}. These are cases, when questions do not apply and are left blank due to filtering procedures \citep[155]{deleeuwPreventionTreatmentItem2003}. The processes generating these types of missing data typically differ. Missings completely at random can be created for example by overlooking a question or a processing error while missings by design is a conscious and deliberate choice in questionnaire design.

In this paper, I focus on missingness due to nonsubstantive answers given by respondents (\textit{item nonresponse}), specifically causes for answer refusals and "Don't know" (DK) answers on the respondent and interview level. Besides unit nonresponse and interview breakoffs, item nonresponse is one of the biggest sources of missing at random and not missing at random. Item nonresponse can also be seen as an indicator for overall data quality since it can indicate satisficing behaviour of the respondent \citep{krosnickResponseStrategiesCoping1991}.

The question-answer process involves multiple steps on behalf of the respondent in the following order: understanding the question, recalling the necessary information from memory, eventually judging information recalled, eventually editing it if the perceived sensitivity or social desirability is high, eventually formating it to fit a response category and finally communicating the response. In all these stages, refusals and DKs can be introduced \citep[pp. 159f]{deleeuwPreventionTreatmentItem2003}.

Refusals and DKs can be attributed to slightly different missingness generating mechanisms though. Refusals are more likely to stem from the editing step when respondents feel an item to be too sensitive. DK is more likely to come about in one of the previous steps if the mental effort to answer a question is too high. But higher mental effort is also related to more refusals \citep{shoemakerItemNonresponseDistinguishing2002}. DK can therefore be seen as a strategy of \textit{(strong) satisficing}, giving a satisfactory answer instead of bothering to actually think about a question in depth (\textit{optimising}) \citep[219]{krosnickResponseStrategiesCoping1991}. Consequentially, DK answers are more likely to be missing at random and refusals are more likely to be not missing at random.

Although the respondent is the one giving the answers in the end, the causes and amplifiers of item nonresponse are manifold and not only vary by which step of the question-answer process they influence \citep{holbrookImpactQuestionRespondent2006} on but also their hierarchical level. This starts with country and survey characteristics and spanning interviewer effects, respondent characteristics, the interview situation, the questionnaire and ends at the individual item. I proceed with a short review of influencing factors of item nonresponse ordered by their level.

\bigskip

\textbf{Country}:
Considerable country-level differences have been observed but there are no convincing explanations for these differences to date. \citet{meitingerPowerCultureItem2020} study how cultural and socio-economic characteristics influence item nonresponse but only uncertainty avoidance turned out to significantly increase levels of item nonresponse. Other country-level variables like individualisation and inequality had the expected direction though. Norms vary geographically that define how sensitive a topic is and what socially acceptable answers are (see below). Some variance between countries stems from composition effects \citep{kochItemNonresponseEuropean2009}.

\textbf{Survey}:
Survey characteristics like mode, topic and incentives as well as instructions to interviewers on refusal conversion, contacting etc. could influence the composition of realised interviews and therefore indirectly have an effect on item nonresponse. Parts of the differences in item nonresponse rates between mail and web modes can be attributed to different demographic compositions of the obtained samples \citep{messerDeterminantsItemNonresponse2012} and refusal conversions often show higher item nonresponse \citep{yanRelationUnitNonresponse2010}. Survey mode also influences the design possibilities of the questionnaire (see below) and the likelihood of respondents or interviewers overlooking questions or processing errors.

\textbf{Interviewer}:
There is often substantial variance in item nonresponse between interviewers. But associations between item nonresponse and interviewer characteristics are not clear cut (e.g. \cite{pickeryImpactRespondentInterviewer1998, pickeryExplorationQuestionCharacteristics2001, silberEffectsQuestionRespondent2021}). But there are hints that older interviewers obtain more item nonresponse and interviewer training can reduce nonresponse \citep{silberEffectsQuestionRespondent2021}.

Explicitly coded refusals or ``Don't know'' answers can also be accidentally misreported by the interviewer or even deliberately changed from a substantial answer to shorten on follow-up questions \citep[165]{deleeuwPreventionTreatmentItem2003}.

\textbf{Respondent}:
The question-answer process requires motivation on behalf of the respondent to cooperate and the mental ability to perform the necessary tasks to get to an answer. Differences in mental ability have quite frequently be related to item nonresponse and are a reasonable explanation for the often observed higher item nonresponse rates of elderly and lower educated respondents \citep{pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}. This reasoning is supported by correlations between item nonresponse and direct measures of intelligence but also to levels of conscentiousness \citep{hedengrenDogThatDidn2012} and to measures of physical and mental health \citep{colsherDataQualityAge1989}. Women and ethnic minorities are also often found to have more item nonresponse \citep{pickeryImpactRespondentInterviewer1998, kupekDeterminantsItemNonresponse1998}. For the latter, this might be connected to the finding that lower literacy and language ability increases item nonresponse \citep{kupekDeterminantsItemNonresponse1998}. \citet{meitingerPowerCultureItem2020} interpret these findings in a social marginalisation and power perspective: item nonresponse reflects broader social inequalities in abilities and access to information.
Respondents that are more interested in the topic of the survey have found to have lower item nonresponse likely due to their higher motivation \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}. Reluctancy to answer and overall skepticism towards surveys and science, privacy concerns, generalised mistrust are personality traits that result in more nonresponse \citep{silberEffectsQuestionRespondent2021}. These are also related to unit nonresponse. The idea of a trade-off between unit and item nonresponse is summarised by \citet{yanRelationUnitNonresponse2010} in their \textit{response continuum model}.

\textbf{Interview}:
Interview characteristics and interactions between interviewer and respondent characteristics have gained a little attention in past research. \citet{vercruyssenEffectSociodemographicMis2017} find less item nonresponse when matching age of interviewers and respondents in the ESS in Belgium while gender matching reduces item nonresponse for males and increases it for females. Matching education has no significant effect after controlling for respondents education. \citet{silberEffectsQuestionRespondent2021} obtain similar results in Germany. Gender matching is also less important while age and education matching are potent predictors of item nonresponse in a sex survey in Taiwan \citep{tuSocialDistanceRespondent2007}.

 Other people present during the interview resulted in higher item nonresponse sometimes \citep{kupekDeterminantsItemNonresponse1998} but not always \citep{tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}. Respondent cooperation as perceived by the interviewer also predicts lower item nonresponse \citep{tuSocialDistanceRespondent2007} but again with exeptions \citep{silberEffectsQuestionRespondent2021}.

\textbf{Questionnaire}:
 Questionnaire design choices like wether DK was explicitly offered but also the complexity of a questionnaire with changes in response categories, routing and filtering are associated with higher nonresponse \citep{messerDeterminantsItemNonresponse2012} likely due to higher likelihoods of mistakes by respondents or interviewers and processing errors. Page layout and progress bars in web surveys have also been discussed to influence item nonresponse \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}. Also, a long questionnaire might fatigue respondents giving more nonsubstantial answers later on.

\textbf{Question}:
The previous point directly relates to a questions position within the questionnaire. But besides fatigue later on, there might also be a learning effect reducing item nonresponse over time \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.
Item non-response varies between questions with more sensitive questions typically having higher item nonresponse. Most prominently, this affects questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998} but \citet{olsonEffectsRespondentQuestion2019} suggests that eventually, a nonsubstantial answer might be even more telling resulting in lower levels of nonresponse. More difficult or unclear questions typically have higher item nonresponse as well but question length of itself is not \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}. Consequentially, attitudinal and behavioural questions often produce more item nonresponse than demographic questions \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}. And design decisions on the question level infuence the cognitive effort needed to provide a substantial answer as well: open-ended questions, being allowed to select multiple options, more response categories and ordering categories are associated with higher nonresponse due to difficulty and burden \citep{holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}. \citet{holbrookInterviewerErrorsHelp2016} found that questions using showcards had more problems although this is likely not because of the showcards but because they are only used with more difficult questions. The question level is one of the best understood levels and constitutes the largest part of the variance. This highlights the importance of the single question for overall data quality \citep{olsonEffectsRespondentQuestion2019}.

\bigskip

To improve data quality, there are a few things we have some control over, mainly considering the interview situation, the interviewers and the survey characteristics including questionnaire and question design. It is therefore reasonable to have a special interest in these to improve data quality. Therefore, the high proportion of explained variance on the question level is good news for survey research. But unfortunately, all of the other manipulatable levels are underresearched. This is most likely due to limited availability of data on the respective concepts.

This article tries to shed some more light on the interview level using paradata and interviewer observations. But since we cannot statistically separate the interview and the unchangeable respondent characteristics, I control for them as well.

\section{Data and Methods}

I analyse item nonresponse in the \textit{European Social Survey (ESS)} Round 9 (version 3.1) \citep{essericEuropeanSocialSurvey2019}. My dependent variables are the number of refusals and DKs as well as a sum of both over all questions that are not affected by routing.

Since our dependent variables are count data and show overdispersion, I opted for a negative binomial regression (NB II) with interviewer fixed effects \citep{allisonFixedEffectsNegativeBinomial2002}. Country differences are absorbed by the interviewer fixed effects as well. I opted for standard errors clustered by interviewer. Since the population of interest in this case are the realised interviews and not societies, no weighting is applied.

Only complete cases are analysed, reducing the sample size from 49,519 to 41,205. The large drop in sample size largely originates from the age of the interviewer and the interview duration, not from item nonresponse on behalf of the respondents. Time stamps have been troublesome in some countries due to errors in the CAPI instrument and Romania used the wrong format for interviewer age \citep{essericESS9CodebookEd2021}. Additionally, some interview durations are suspiciously long, so I top-coded the length of the interview to 3 hours.

The respondent-level variables are from the original questionnaire while the interview-level variables are either calculated from paradata or are interviewer observations recorded after the interview in the interviewer questionnaire. Most predictors are measured on a 5-point Likert scale. If they are not, they are either recoded to make relative effect size comparisons or their scale is mentioned in brackets in the results section. Categorical predictors (gender, group memberships, interview interference, interview language, gender and age matching) are included as dummy variables.

The analyses are carried out in \textit{R} \citep{rcoreteamLanguageEnvironmentStatistical2021} using the \textit{Tidyverse} packages \citep{wickhamWelcomeTidyverse2019} for data handling and graphics and the \textit{fixest} package \citep{bergeEfficientEstimationMaximum2018} to estimate the regression. The code is available \href{https://github.com/maltegroenemann/inr_ess9}{here}.

\section{Results}

\begin{figure}
\caption{Results of the Fixed-Effects Negative Binomial Regression}
 \includegraphics[width=.99\textwidth]{results.pdf}
 \label{results}
\end{figure}

Figure \ref{results} shows the regression results for the three dependent variables. Starting with the respondent level, older people are associated with slightly less DKs but more refusals leading to no significant relation between age and total item nonresponse. Women show slightly more DKs than men. Higher education is associated with less DKs but more refusals. Members of ethnic minorities are associated with more DKs while members of discriminated groups have more refusals on average. The more respondents are interested in politics, the fewer DKs and refusals they give in this mainly political survey. The more respondents believe that the political system allows them to have influence, the fewer DKs. A better health is associated with fewer DKs but more refusals. And finally, the more respondents trust generalised others, the fewer DKs and refusals.

At the interview level, a longer interview is associated with more of both types of nonresponse. But other people present or interfering during the interview shows no relation to either. If the interview is conducted in a language different from the one primarily spoken at the respondents home, more DKs and refusals are predicted. Matching genders of the interviewer and the respondent has no relation to item nonresponse and remains so when interacted with respondents gender (not shown). More DKs are predicted when the respondent is more than ten years older than the interviewer. The extent of the respondents accurately using the showcards is negatively related to both types of nonresponse. The respondent more often asking for clarifications is associated with more DK nonresponse. The interviewer perceiving the respondent to be reluctant increases both DKs and especially refusals. The more repondents tried to answer to the best of their abilities, the fewer nonresponse and the better the respondents understands the questions, the fewer DKs.

If not mentioned otherwise, the relations concerning the total amount of item nonresponse mirror the relation with DK. This is hardly surprising since DK is much more frequent. There are in total 176,490 DKs (on average 3.5 per respondent) over all variables unaffected by routing but only 41,821 refusals (0.8 per respondent).

The results from Figure \ref{results} are displayed in a regression table in the appendix as well.

\section{Conclusion}

This article set out to improve our understanding of item nonresponse, a typical indicator of data quality, focussing on the interview- and the respondent-level.

Compared to previous results and common explanations for repondent-level variables, most results are in line with them: better educated, politically more interested and engaged, more trusting male ethnic majority respondents have lowest rates of item nonresponse. But there are some surprising findings as well: education differs in its effect by type of nonresponse. This might be due to them being more able and knowing but at the the same time more conscious about privacy. Privacy concerns might also be hightened for members of discriminated groups resulting in more refusals. Health did also differ in its effect by type of nonresponse. Age on the other hand has not the expected effect though. But this is likely due to controlling for measures of cognitive ability. This holds true for all variables though, the effects are all conditional on the other independent variables. Understanding their separate effects and potential mediations is beyond the scope of the article though.

The more interesting results however are the ones on the interview level because they can actually be changed and therefore be considered in the design of future surveys. Interference of the interview and matching respondents and interviewers gender has no effect at least in this attitude survey. With more sensitive topics, this might be different though. A longer interview and asking for many clarifications are associated with more item nonresponse but are likely just resulting from the common cause of understanding difficulties. Taken together with the interviewers assessment of the respondents understanding of the questions, this again highlights the importance of question design for data quality. One possibility to improve understanding is to conduct the interview in the respondents primary language but it is a trade with costs and comparability. Increasing respondents motivation, the underlying concept of whether respondents answered to their best ability, also could improve data quality. But besides understanding of the questions, the effects largest in magnitude are respondents reluctancy (especially considering refusals which are likely to be not missing at random) and their use of showcards. Building a trusting interview situation and making sure that respondents correctly use the showcards are the most promising avenues to reduce item nonresponse.

But as with all cross-sectional analyses, these coefficients merely represent associations, not causal effects. Separating the causal effects using survey experiments as well as improving on our understanding of mediating and confounding relations between these variables are left for further research. Especially since most interview-level predictors are subjective assessments by the interviewers and therefore endogenous \citep[8]{silberEffectsQuestionRespondent2021}. And there might still be variables and concepts on both levels that have been omitted due to inavailability. Examples are attitudes on personal privacy, trust in science (coming in ESS10) as well as better measures of motivation and commitment.

\printbibliography

\section*{Appendix: Table of the Regression Results}
\small
\input{results.tex}
\end{document}

### 2021-08-25 inr_ESS9.R

@@ -1,0 +1,125 @@
+###########################################################################
+### Item Nonresponse in ESS Round 9
+### Author: Malte Grönemann
+### Version: R 4.0.5 on Linux
+###########################################################################
+
+library(fixest)
+library(tidyverse)
+library(haven)
+
+
+# import of the raw data and joining
+# necessary to change encoding on Linux/OSX, remove if on Windows. 
+# See https://haven.tidyverse.org/reference/read_dta.html#character-encoding
+ESS9 <- full_join(read_dta("ESS9e03_1.dta", encoding = "latin1"), 
+                  read_dta("ESS9INTe03.dta", encoding = "latin1"), 
+                  by = c("idno", "cntry"))
+
+# calculating the numbers of refusals and DK by interview
+# See https://haven.tidyverse.org/articles/semantics.html#tagged-missing-values-1
+# ESS coding: Not applicable -> a, Refusal -> b Don't know -> c, Not available -> d
+count_refusals <- ESS9 %>% 
+  select(nwspol:impfun) %>% # only substantial variables
+  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>% # excluding vars affected by routing
+  mutate(across(.fns = ~ is_tagged_na(.x, tag = "b"))) %>%
+  rowwise() %>%
+  transmute(n_refusals = sum(c_across()))
+
+count_dk <- ESS9 %>% 
+  select(nwspol:impfun) %>%
+  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>%
+  mutate(across(.fns = ~ is_tagged_na(.x, tag = "c"))) %>%
+  rowwise() %>%
+  transmute(n_dk = sum(c_across()))
+
+# total numbers of DK and REF
+sum(count_dk$n_dk)
+sum(count_refusals$n_refusals)
+
+# selecting and if necessary transforming variables for analysis
+analysis <- ESS9 %>%
+  transmute(`Age (in 10 years)` = agea / 10, # respondent
+            `Female` = gndr - 1,
+            `Education (ISCED)` = ifelse(eisced < 10, eisced, NA),
+            `Ethnic Minority Member` = blgetmg * -1 + 2,
+            `Member of discriminated Group` = dscrgrp * -1 + 2,
+            `Political Interest` = (polintr-1) * -4/3 + 5,
+            `People have Political Influence` = psppipla,
+            `Health` = health * -1 + 6,
+            `Generalised Trust` = ppltrst/2,
[…]

### 2022-07-04 inr_ESS9_supp.Rmd
Version under Review at SRM

---
title: "Supplement to 'How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey'"
author: "Malte Grönemann, University of Mannheim"
date: "04/07/2022"
output: pdf_document
---

```{r settings, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
options(digits = 3)
```

# Version and Packages

This analysis was written for `r R.Version()$version.string` on Linux. The following packages were used with the respective versions as comments.

```{r setup}
library(tidyverse) # 1.3.1
library(haven) # 2.5.0
library(qqplotr) # 0.0.5
library(fixest) # 0.10.4
```

# Data Manipulation

`ESS9e03_1.dta` is the main data file from the ESS 9 version 3.1 and `ESS9INTe03.dta` is the interviewer questionnaire. The data are available after registration here: https://www.europeansocialsurvey.org/data/download.html?r=9
*idno* is a respondent identifier and *cntry* is the country abbreviation.

It is necessary to change the encoding on Linux/Mac to *latin1*, remove the option if on Windows.
See https://haven.tidyverse.org/reference/read_dta.html#character-encoding

```{r read_data}
ESS9 <- full_join(read_dta("ESS9e03_1.dta", encoding = "latin1"), 
                  read_dta("ESS9INTe03.dta", encoding = "latin1"), 
                  by = c("idno", "cntry"))
```

## Dependent Variables

I use three dependent variables: the number of refusals by a respondent, the number of Don't know (DK) and their sum. I sum them over all questions applicable to all respondents. That means, all questions only asked a subset of respondents due to previous answers or survey experiments are left out.

As background information, refusal and DK are not read out to the respondents in the ESS or on the showcards. But interviewers have them as distinct options within their CAPI instrument. Interviewers are advised to accept them without probe. Whether the respondent refuses or says DK is therefore an interpretation of the interviewer.

`select(nwspol:impfun)` selects only the variables that are based on questions asked the respondent and gets rid of the interviewer questionnaire, weights etc. for the calculation of item nonresponse.

The ESS distinguishes different types of missings using labeled missings in Stata. Their codes are:

- .a - not applicable
- .b - respondent refused
- .c - Don't know
- .d - not available (should not exist: code for processing errors)

The package *haven* has a function to detect these tagged missings in a Stata dataset in R, see https://haven.tidyverse.org/articles/semantics.html#tagged-missing-values-1

To have a comparable set of items for all respondents, I only select the ones applicable to all: `select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE))`

Attention: the step to calculate the number of missings for each respondent takes some time.

```{r dep_var}
count_refusals <- ESS9 %>% 
  select(nwspol:impfun) %>%
  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>%
  mutate(across(.fns = ~ is_tagged_na(.x, tag = "b"))) %>%
  rowwise() %>%
  transmute(n_refusals = sum(c_across()))

count_dk <- ESS9 %>% 
  select(nwspol:impfun) %>%
  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>%
  mutate(across(.fns = ~ is_tagged_na(.x, tag = "c"))) %>%
  rowwise() %>%
  transmute(n_dk = sum(c_across()))

rbind(tibble(n = count_refusals$n_refusals,
             type = "Refusal"),
      tibble(n = count_dk$n_dk,
             type = "DK")) %>%
  ggplot() +
  aes(n) +
  geom_histogram() +
  facet_wrap(~type) +
  labs(title = "Fig. S1: Histogram of INR by Type")
```

There are much more DKs than refusals. There are `r sum(count_dk$n_dk)` DKs, on average `r sum(count_dk$n_dk) / length(count_dk$n_dk)` per respondent. The number of refusals amounts to `r sum(count_refusals$n_refusals)`, on average `r sum(count_refusals$n_refusals)/length(count_refusals$n_refusals)` per respondent. That totals `r sum(count_refusals$n_refusals) + sum(count_dk$n_dk)` cases of item nonresponse. There is a correlation of `r cor(count_dk$n_dk, count_refusals$n_refusals)` between the number of DKs and refusals.

The standard deviation of DK is `r sd(count_dk$n_dk)` and the of refusals is `r sd(count_refusals$n_refusals)`. Although for the regression the variance of the residuals matters, this is a first sign of overdispersion.

# Independent Variables

## Preparation for Analysis

The variables from the interviewer questionnaire still contain missing values as number codes 8 and 9. They are set to NA if their value is above 6. Bulgaria used the wrong format for interviewer age and therefore adds many missings to it and is resultantly excluded. The *other* category of ISCED (number code 55) is coded to NA as well. The length of the interviews are top coded to 3 hours because there were errors with the CAPI instruments time stamps in some occasions (see ESS 9 codebook).

```{r prep}
analysis <- ESS9 %>%
  zap_label() %>% zap_missing() %>% zap_formats() %>% # getting rid of Stata formatting
  transmute(age = agea, # respondent
            age_sq = agea^2,
            educ = ifelse(eisced < 10, eisced, NA),
            duration = ifelse(inwtm < 180, inwtm / 30, 6),
            interference = ifelse(preintf < 6, preintf * -1 + 2, NA),
            notprimlang = as.integer(lnghom1 != intlnga),
            gndrmatch = ifelse(intgndr < 6, as.numeric(intgndr == gndr), NA),
            resolder = ifelse(intagea < 200, as.numeric(agea - intagea > 10), NA),
            intolder = ifelse(intagea < 200, as.numeric(intagea - agea > 10), NA),
            showcards = ifelse(resswcd < 6, -1*resswcd + 4, NA),
            clarif = ifelse(resclq < 6, resrelq, NA),
            bestab = ifelse(resbab < 6, resbab, NA),
            underst = ifelse(resundq < 6, resundq, NA),
            intver = paste0(intnum, cntry), # FE
            n_dk = count_dk$n_dk, # DV
            n_ref = count_refusals$n_refusals,
            n_tot = n_dk + n_ref)

# somehow, I had trouble recoding clarif in transmute
analysis$clarif[analysis$clarif > 6] <- NA

labels <- c(age = "Age",
            age_sq = "Age squared",
            educ = "Education (ISCED)",
            underst = "Understood Questions",
            bestab = "Answered to best Ability",
            duration = "Duration of Interview (30 Min)",
            interference = "Interference of Interview",
            notprimlang = "Int. not in primary Language",
            gndrmatch = "Gender Matching",
            resolder = "Respondent 10 years older",
            intolder = "Interviewer 10 years older",
            showcards = "Use of Showcards",
            clarif = "Amount of Clarifications",
            intver = "Interviewer",
            n_dk = "Don't know",
            n_ref = "Refusal",
            n_tot = "Total")
```
## Descriptive Statistics

### Variables of Interest

Except of the use of showcards and the duration of the interview, all variables of interest are dummys:

The proportion of interviews where the interviewer is more than 10 years older is `r mean(analysis$intolder, na.rm = T)` and the proportion where the respondent is more than 10 years older is `r mean(analysis$resolder, na.rm = T)`. Both are based of the same original variables, age of the respondent and age of the interviewer and have therefore the same number of NAs: `r sum(is.na(analysis$intolder))`. The large number of NAs is predominantly due to missings in interviewers age. It is missing very frequently. Romania used the wrong format for this item and therefore all interviews from Romania are excluded via missingness in these items.

The proportion of interviews where respondent and interviewer were of the same gender is `r mean(analysis$gndrmatch, na.rm = T)` (`r sum(is.na(analysis$intolder))` NAs).

In `r mean(analysis$interference, na.rm = T)*100` percent of the interviews, someone besides the interviewer and respondent was present in the same room or interfered with the interview (`r sum(is.na(analysis$interference))` NAs).

`r mean(analysis$notprimlang, na.rm = T)*100` percent of the interviews were conducted in a language different from the one the respondent primarily speaks at home. There are no missings in this item.

The extent the respondent used showcards as perceived by the interviewer is measured on a three point scale: respondent used all the applicable showcards, respondent used only some applicable showcards, respondent refused/ was unable to use the showcards at all. This question is asked to the interviewer after the interview, see the ESS9 source questionnaire. I flipped the order to ease interpretation: higher number represents more frequent use. Most respondents used all the applicable showcards (`r mean(analysis$showcards == 3, na.rm = T)`), `r mean(analysis$showcards == 2, na.rm = T)` respondents used them only sometimes and `r mean(analysis$showcards == 1, na.rm = T)` never used them. There are `r sum(is.na(analysis$showcards))` NAs.

The duration of the interview is automatically computed by the CAPI instrument based on start and end time stamps. There have been some issues with the time stamps in some countries though, resulting in some unusually long durations. I top-coded them to 3 hours and rescaled the variables to 30 minutes.

```{r duration}
summary(analysis$duration)

ggplot(analysis) +
        aes(x = duration) +
        geom_histogram() +
        labs(x = "Interview duration", y = "")
```

### Control Variables

The ESS surveys European populations over the age of 15. This is therefore the minimum age in the sample. The maximum age is `r max(as.numeric(analysis$age), na.rm = T)`. The mean age is `r mean(as.numeric(analysis$age), na.rm = T)` with a standard deviation of `r sd(as.numeric(analysis$age), na.rm = T)`. There are `r sum(is.na(analysis$age))` NAs.

```{r age}
ggplot(analysis) +
        aes(x = as.numeric(age)) +
        geom_histogram() +
        labs(x = "Age", y = "")
```

Education is operationalised using the established ISCED scale that orders the latest educational degrees by level. The number codes and levels are:

- 1: ES-ISCED I , less than lower secondary
- 2: ES-ISCED II, lower secondary
- 3: ES-ISCED IIIb, lower tier upper secondary
- 4: ES-ISCED IIIa, upper tier upper secondary
- 5: ES-ISCED IV, advanced vocational, sub-degree
- 6: ES-ISCED V1, lower tertiary education, BA level
- 7: ES-ISCED V2, higher tertiary education, >= MA level

`r table(analysis$educ, useNA = "always")`

All control variables from the interviewer questionnaire are measured on a five point scale: Never, Almost never, Now and then, Often, Very often with number codes from 1 to 5.

I selected the assessment of the interviewer how often the respondent...
- asked for clarifications,
- answered to the best of their ability and
- understood the question.

These items have two functions as controls. They are proxies of ability (and motivation) as well as the other controls. And they control for the interviewers' assessment of the interview which might influence the answer on showcard use. In multiple regression, only the effect of a variable *net of all others* is estimated, therefore dampening the endogeneity problem with the showcard use item.

```{r, intvar}
rbind(tibble(value = analysis$clarif,
             var = "Amount of Clarifications"),
      tibble(value = analysis$bestab,
             var = "Answered to best Ability"),
      tibble(value = analysis$underst,
             var = "Understood Questions")) %>%
  ggplot() +
    aes(x = value) +
    geom_bar() +
    facet_wrap(~var)

rbind(tibble(value = analysis$clarif,
             var = "Amount of Clarifications"),
      tibble(value = analysis$bestab,
             var = "Answered to best Ability"),
      tibble(value = analysis$underst,
             var = "Understood Questions")) %>%
  group_by(var) %>%
  summarise(Mean = mean(value, na.rm = TRUE),
            SD = sd(value, na.rm = TRUE))
```

# Multivariate Analyses

I analyse the data using a negative binomial regression with interviewer fixed effects . Standard errors are clustered by interviewer. Since the population of interest in this case are interviews and not countries, no weighting is applied.

The missings are deleted listwise.

```{r model}
model <- fenegbin(data = analysis,
                  fml =  c(n_dk, n_ref, n_tot) ~
                    duration + interference + showcards + notprimlang +
                      gndrmatch + resolder + intolder +
                      educ + age + age_sq + underst + bestab +  clarif |
                      intver,
                  se = "cluster")

setFixest_dict(labels)
etable(model, 
       file = "results_inr_ess9.tex", replace = TRUE, tex = TRUE,
       title = "Regression Results",
       digits = 3, se.below = TRUE,
       signifCode = c("***"=0.001, "**"=0.01, "*"=0.05),
       fitstat = c("n", "pr2", "wpr2", "bic", "theta", "f"))
etable(model,
       tex = FALSE,
       title = "Regression Results",
       digits = 3, se.below = TRUE,
       signifCode = c("***"=0.001, "**"=0.01, "*"=0.05),
       fitstat = c("n", "pr2", "wpr2", "bic", "theta", "f"))

results <- bind_rows(bind_cols(var = rownames(model$n_dk$coeftable), 
                               model$n_dk$coeftable, 
                               y = "Don't know"),
                     bind_cols(var = rownames(model$n_ref$coeftable),
                               model$n_ref$coeftable, 
                               y = "Refusal"),
                     bind_cols(var = rownames(model$n_tot$coeftable),
                               model$n_tot$coeftable, 
                               y = "Total")) %>%
  rename(estimate = `Estimate`,
         se = `Std. Error`) %>%
  filter(var != ".theta") %>%
  filter(var %in% c("duration", "interference", "showcards", "notprimlang",
                    "gndrmatch", "resolder", "intolder")) %>%
  mutate(irr = exp(estimate),
         upper = exp(estimate + se * (-qnorm((1-0.95)/2))),
         lower = exp(estimate - se * (-qnorm((1-0.95)/2))))

ggplot(results) +
  aes(y = reorder(var, -irr),
      x = irr,
      colour = y,
      shape = y) +
  geom_vline(xintercept = 1, 
             colour = "darkgrey") +
  geom_point(size = 2) +
  geom_linerange(aes(y = var, 
                     xmin = lower, 
                     xmax = upper),
                 lwd = .5) +
  facet_wrap(~factor(y, levels=c("Total", "Don't know", "Refusal"))) +
  scale_color_manual(values = c("#E69F00", "#009E73", "#D55E00")) +
  scale_shape_manual(values = 15:17) +
  scale_y_discrete(labels = labels) +
  theme(legend.position = "none", 
        strip.background = element_blank(),
        text = element_text(size = 15)) +
  labs(y = "", x = "Incidence Rate Ratios with 95% Confidence Intervals")
ggsave(filename = "results_inr_ess9.pdf",
       width = 22, height = 10, units = "cm")
```

Although I removed all NAs before, there are different numbers of observations for each dependent variable. This is due to some interviewers having no within-variance (that have only respondents with not a single item nonresponse).

## Regression Diagnostics and Model Justification

Fixed effects count data models are quite contested. Wooldridge (1999) claims that fixed effects negative binomial regression (FENB) is not a true fixed effects estimator and that overdispersion is not of much concern in these models. He therefore advises to always use FE Poisson regression (FEPois) over FENB.

Allison and Waterman (2002) agree that FENB in its previous form was not a true FE estimator but propose a modified version that they advise to always use over FEPois. This improved version is the one implemented in the fixest package.

I compare estimates and diagnostic plots for both regression models using the total sum of item nonresponse as the dependent variable to assess reliability of my results.

```{r diag}
fepois <- fepois(data = analysis,
                  fml =  n_tot ~
                    duration + interference + showcards + notprimlang +
                      gndrmatch + resolder + intolder +
                      educ + age + age_sq + underst + bestab +  clarif |
                      intver,
                  se = "cluster")

setFixest_dict(labels)
etable(model$n_tot, fepois,
       tex = FALSE,
       title = "Tab. S2: FENB vs. FEPois",
       digits = 3, se.below = TRUE,
       signifCode = c("***"=0.001, "**"=0.01, "*"=0.05),
       fitstat = c("n", "pr2", "wpr2", "bic", "theta", "f"))

diagnostics <- rbind(tibble(resid = resid(fepois,
                                          type = "deviance"),
                            fitval = fepois$fitted.values,
                            model = "FEPois"),
                     tibble(resid = resid(model$n_tot,
                                          type = "deviance"),
                            fitval = model$n_tot$fitted.values,
                            model = "FENB"))

ggplot(diagnostics) +
  aes(sample = resid) +
  stat_qq(size = 0.1) +
  stat_qq_line() +
  stat_qq_band() +
  facet_wrap(~model) +
  labs(title = "Fig. S2: QQ-Plot of Deviance Residuals",
       x = "", y = "")

ggplot(diagnostics) +
  aes(x = fitval,
      y = resid) +
  geom_point(size = 0.1) +
  facet_wrap(~model) +
  labs(title = "Fig. S3: Residuals vs Fitted: FENB",
       x = "Fitted Values",
       y = "Deviance Residuals")
```

As can be seen in the diagnostic plots, the Poisson Regression with fixed effects exhibits a much larger residual variance and the deviance residuals are much less normal than the negative binomial regression with fixed effects. The average deviance residual of the FEPois is `r mean(resid(fepois, type = "deviance"))` with a standard deviation of `r sd(resid(fepois, type = "deviance"))` compared to `r mean(resid(model$n_tot, type = "deviance"))` with a standard deviation of `r sd(resid(model$n_tot, type = "deviance"))` for FENB.

Although normality of the residuals is not such a strong criterion with count data models, Cameron and Trivedi (2013, chapter 9.4) expect normality in deviance residuals in correctly specified models. While not perfect, the deviance residuals of the FENB model are much closer to normally distributed than FEPois. And the unconditional FENB by Allison and Waterman (2002) implemented in the fixest package is a true fixed effects model unlike previous versions. It has an incidental parameter problem though, but it is usually very small. Wooldridge (1999) argued for Poisson regression when using fixed effects but this seems to be the worse option in this case. The overdispersion seems to be still consequential even with fixed effects.

On the other hand, the substantial results are very similar. Coefficients differ slightly between models but signs and significance are largely identical. I therefore settle on the FENB model in the paper.

Clustered standard errors are the default in the fixest package. Normal parametric standard errors tend to be way too large (Cameron and Trivedi 2013). I replicated the regressions in Stata using bootstrap standard errors (not shown) and only get marginally different results, likely from differences in the optimisation algorithm.

# Literature

Allison, Paul D. and Richard P. Waterman (2002). “Fixed-Effects Negative Binomial Regression Models”. In: Sociological Methodology 32.1, pp. 247–265. doi: 10.1111/14679531.00117.

Cameron, A. Colin, and Pravin Trivedi. Regression Analysis of Count Data. 2nd ed. Cambridge: Cambridge University Press, 2013. https://doi.org/10.1017/CBO9781139013567.

Wooldridge, Jeffrey M. (1999). “Distribution-Free Estimation of Some Nonlinear Panel Data Models”. In: Journal of Econometrics 90.1, pp. 77–97. doi: 10.1016/S0304-4076(98)00033-5.

### 2022-07-04 inr_ESS9_text.tex
Version under Review at SRM

\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage[UKenglish]{babel}
\usepackage[babel]{csquotes}

\usepackage{amsmath, amssymb}
\usepackage{graphicx, subcaption, booktabs}

\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[all]{hypcap}

\usepackage[style=authoryear, backend=biber, natbib]{biblatex}
\bibliography{~/Zotero/Bibliothek.bib}

\usepackage{setspace}
\doublespacing

%opening
\title{How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey}
\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}

\begin{document}

\maketitle

\begin{abstract}
I review the literature on item nonresponse in surveys and based on this review, I extend the satisficing model with respondents' privacy concerns to incorporate all relevant aspects of the response process for item nonresponse. I review proposed strategies to reduce item nonresponse and test selected strategies. Results suggest that boosting respondents' use of showcards, shorter interview durations and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse. Other people present during the interview only has a small association with the number of refusals. Matching age and gender of respondents and interviewers appears not to be a worthwile strategy.
\end{abstract}

Keywords: item nonresponse, survey design, survey error, face-to-face surveys, fixed-effects negative binomial regression

\section{Introduction}

Missing data pose a problem to the analysis of survey data. They decrease the effective sample size and can introduce bias to estimates if the causes for missingness are related to the item or respondent characteristics \citep{deleeuwPreventionTreatmentItem2003}. Although missing data are relatively rare in most single items, they can add up to a considerable loss of observations in multivariate analyses. Due to these harmful effects of missing data, one objective of survey researchers is to keep their prevalence as low as possible. Therefore, it is important to understand the processes that can lead to missing data.

I focus on the respondent as a source of missing data by answering with ``Don't know'' (DK) or refusing to answer. (\textit{item nonresponse}). Item nonresponse can also be seen as an indicator of overall data quality since it can be the result of \textit{satisficing} \citep{krosnickResponseStrategiesCoping1991}. Satisficing means that the respondent is giving a satisfactory answer instead of the best one possible.

I review the existing literature on item nonresponse and extend the satisficing model based on this review to include privacy concerns resulting in an encompassing model of item nonresponse: the probability of item nonresponse depends on the task difficulty of the item(s) divided by the product of ability and motivation of the respondent or the respondents' privacy concerns. This means that higher difficulty results in more item nonresponse, higher ability and motivation in less item nonresponse. Difficulty, ability and motivation are separate from privacy concerns, e.g. due to item sensitivity or general mistrust. Privacy concerns are relevant when deciding whether to disclose an already formed answer and does not influence the cognitive burden of retrieving the answer.

In the empirical part of the paper, I explore a few practical ideas to reduce item nonresponse. Promoting the use of showcards, shorter interviews and tranlating questionnaires might be promising strategies. All of those reduce the cognitive burden of respondents. The effects are not causal though because I am not able to sufficiently control for respondent ability. Matching respondents' and interviewers' gender and age seem not to be efficient tools to reduce item nonresponse. They could have influenced item nonresponse if respondents are more willing to share private information with interviewers that are similar to them.

This introduction follows a short theoretical discussion presenting a model for the probability of item nonresponse based on satisficing. Structured by this model, I review the literature on strategies how to reduce item nonresponse and test a selection using the \textit{European Social Survey}. The rest of the paper is devoted to this test of strategies, describing data and methods and presenting results. Finally, I summarise and discuss my review and results.

\section{A Model of Item Nonresponse}

The \textit{survey response process} \citep{tourangeauPsychologySurveyResponse2000} involves multiple steps on behalf of the respondent. They need to comprehend the question, retrieve information from memory, eventually they need to judge these information, map them onto the response options and perhaps edit the response due to sensitivity or social desirability. Respondents can take these steps in order or jump back and forth, e.g. if they need to form an opinion on the spot. But in all of these steps, item nonresponse can be introduced \citep{deleeuwPreventionTreatmentItem2003}.

The two types of item nonresponse might be related to different steps of the survey response process. Refusals are more likely to stem from the editing step when respondents do not want to answer although they could. They might feel an item is too sensitive or do not trust the interviewer with this information. DK is more likely to come about when the respondent cannot answer in the first place. Either the respondent cannot answer because they do not know about the content of the question or are unable to remember an event \citep{beattyAnswerNotAnwer2002, turnerCanResponseLatencies2015}. Or they do not perceive it worth the effort to come up with an answer and satisfice \citep{krosnickResponseStrategiesCoping1991}. But \citet{shoemakerItemNonresponseDistinguishing2002} have shown that higher mental effort is related to more refusals as well. And conversely, it is plausible that DK is used as a more polite way to refuse. I, therefore, do not distinguish between them in the following sections even though they often show different strengths of predictors \citep{silberEffectsQuestionRespondent2021}.

Ability has been the explanation for the differences often found in item nonresponse by education, age, physical and mental health \citep{colsherDataQualityAge1989, pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}. Women and ethnic minority members are often found to have more item nonresponse \citep{kupekDeterminantsItemNonresponse1998, pickeryImpactRespondentInterviewer1998}. For the latter, on average lower literacy and lower command of the majority language are hypothesised to cause more item nonresponse \citep{kupekDeterminantsItemNonresponse1998}. \citet{meitingerPowerCultureItem2020} conclude that item nonresponse reflects broader social inequalities in abilities and access to information. The ability hypothesis is also directly supported by correlations between item nonresponse and measures of intelligence \citep{hedengrenDogThatDidn2012}.

More difficult or unclear questions typically have higher levels of item nonresponse \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}. Since demographic questions can often be recalled very simply by respondents, they often show less item nonresponse than attitudinal and behavioural questions where the answers often do not exist before before getting asked the question \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.

Even when able to perform the task at hand, respondents not necessarily want to.
They need to perceive the effort as worth it. Differences in \textit{motivation} are an explanation
for less item nonresponse of respondents that are more interested in the topic of the survey \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}. Item nonresponse is related to measures of conscientiousness as well \citep{hedengrenDogThatDidn2012}.

Motivation can also affect the editing process. For example, respondents (and interviewers) might deliberately not answer screening and filtering questions to reduce survey burden \citep{tourangeauMotivatedMisreportingShaping2015}.
When editing an answer, concerns of privacy\footnote{I use this label to subsume general concerns about privacy towards the survey (e.g. about data processing, anonymity) and not wanting to answer single sensitive questions.} come into play. Am I going to be judged by the interviewer when answering honestly? Are my data really secure and confidential? This is a particular problem for questions perceived as more intrusive \citep{tourangeauSensitiveQuestionsSurveys2007}. Most prominently, this affects questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998}, which often show particularly high levels of item nonresponse. High item nonresponse is often used as an indicator for question sensitivity \citep{tourangeauSensitiveQuestionsSurveys2007}. Reluctancy and overall skepticism towards surveys and science, privacy concerns and generalised mistrust are traits that result in more nonresponse \citep{silberEffectsQuestionRespondent2021}.

\bigskip

I have thus identified four constructs related to the probability of item nonresponse under which most research can be subsumed: cognitive ability, task difficulty, motivation and privacy concerns. However, as \citet{krosnickResponseStrategiesCoping1991} already hypothesised, these concepts are interrelated in their effect on item nonresponse. Very easy questions can be answered by less able respondents and very hard questions might even cause the most able to struggle. The resulting fraction of difficulty by ability represents the relative mental effort to answer a question. To what point the relative mental effort is worth it depends on the respondents' motivation. A highly motivated respondent even answeres the difficult questions. The probability of nonresponse decreases with motivation. \citet[225]{krosnickResponseStrategiesCoping1991} has summarised these relations into a formula of the probability of satisficing. Satisficing refers to respondents lowering their standards in answering a question and only giving a satisfactory instead of an optimal answer to cut down on cognitive effort. Item nonresponse is such a satisficing strategy. There are situations where DKs are genuine though \citep{turnerCanResponseLatencies2015}. This is nonetheless a product of the respondents' ability and thereby included in the formula.

Additionally, higher privacy concerns lead to more item nonresponse. Since this relates to another step in the survey response process, namely editing rather than comprehension, retrieval or judging, I postulate it to be independent from the other concepts. I therefore suggest extending the formula by Krosnick as follows:

$$ P_{INR} \ = \ max \left( \frac{difficulty}{ability \ * \ motivation} \ , \ privacy \right) $$

\textit{The probability of a nonsubstantive answer by the respondent is given by the task difficulty
divided by the product of ability and motivation or the respondents' privacy concerns, whichever
is higher.}

This formula specifically addresses item nonresponse for a single item but its meaning is adaptable to every level of a survey.

\section{How to reduce Item Nonresponse}

Based on these four concepts, one can theorise how to decrease item nonresponse. Some of the following possibilities may seem obvious and are established standards not only because of their potential relationship to item nonresponse but to ensure quality of substantial answers as well. Others might reduce item nonresponse but they could have negative consequences for other parts of \textit{total survey error}, the combined effect of all error sources in a survey \citep{grovesTotalSurveyError2010}. They require a trade-off before implementation.

I have structured this review of approaches to reduce item nonresponse by the respective concepts they target.

\subsection{Task Difficulty}

Task difficulty is predominantly influenced by design choices regarding single questions and their type of expected response. For the design of questions, much typical advice from question design aims to reduce task difficulty: asking short questions with simple structures, avoiding ambiguity and fuzzy concepts and not asking respondents to do multiple things at once. For a more comprehensive review on advice on question and questionnaire design, see for example \citet{smythDesigningQuestionsQuestionnaires2016}. But the type of question is also associated with task difficulty: open-ended questions, being allowed to select multiple options, more response categories and being asked to order categories are associated with higher item nonresponse \citep{schumanOpenClosedQuestion1979, holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}. Additional measures to decrease difficulty can be visual helpers like images or showcards. With showcards, respondents do not need to remember all response categories while thinking about the question. But there is not much empirical work on the effect of showcards on item nonresponse. \citet{holbrookInterviewerErrorsHelp2016} found that questions using showcards had more item nonresponse although this is likely because they are only used with more difficult questions. In the context of the ESS, showcards do not seem to affect the distribution of substantive responses in experiments \citep{jackleAssessingEffectData2010} but item nonresponse was not investigated.

The question level is very well understood and constitutes the largest part of the variance in item nonresponse. This highlights the importance of the single question for overall data quality \citep{olsonEffectsRespondentQuestion2019}.

But not only single questions but the whole questionnaire needs to be designed as simply
as possible. Questionnaires with changes in response categories, routing and filtering are associated with higher nonresponse \citep{messerDeterminantsItemNonresponse2012}. Grouping questions by topic could reduce mental effort and therefore item nonresponse but it also increases the likelihood of non-differentiation between the items \citep{krosnickResponseStrategiesCoping1991}. Offering DK and refusal explicitly increases their use because respondents are more aware of the “easy way out” \citep{schumanAssessmentNoOpinion1979, beattyAnswerNotAnwer2002}.

Task difficulty can be reduced for language minorities by providing the questionnaire in more languages. However, this is expensive and can reduce comparability of cases. For a review on comparability in cross-cultural surveys, see e.g. \citet{behrTranslationMeasurementInstruments2016}.

\subsection{Ability}

To make sure that the respondent is able to answer the questions, interviews should be conducted in situations where understanding and concentration is as good as possible. Other people present during the interview, for example, could be a source of distraction. Empirical investigations have not been conclusive though \citep{kupekDeterminantsItemNonresponse1998, tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}. With longer interviews, respondents might get tired \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.

\subsection{Motivation}

Motivation could decrease over the course of the interview as well. Different page layouts and progress bars might help with that in web surveys but evidence is mixed \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}. The
literature on cooperation enhancement, e.g. using incentives, has focussed on unit nonresponse. Some of these ideas might be applicable to increase respondent motivation within the interview as well. After all, unit and item nonresponse are linked: respondents that initially refused to participate have higher levels of item nonresponse \citep{yanRelationUnitNonresponse2010, frickerExaminingRelationshipNonresponse2010}.

\subsection{Privacy Concerns}

To tackle reluctancy to answer due to privacy concerns, survey researchers need to make sure that the respondents' data are secure and anonymous and communicate that. Explaining why the data are collected, how they are processed and what is done to protect their privacy is also ethically advisable and often legally required.

In face-to-face surveys, establishing trust depends on the respondent-interviewer dynamics. Based on the hypothesis that respondents might perceive interviewers similar to themselves as more trustworthy, scholars have assessed the impact of matching their characteristics. \citet{vercruyssenEffectSociodemographicMis2017} find less item nonresponse when matching age
of interviewers and respondents and matching gender reduces item nonresponse for males and increases it for females. Matching education has no effect \citep{vercruyssenEffectSociodemographicMis2017, silberEffectsQuestionRespondent2021}. \citet{tuSocialDistanceRespondent2007} on the other hand find age and education matching are potent predictors of item nonresponse.

\section{Strategies to be Tested}

In this review, I have stressed the role of the respondent and how design choices can increase the probability of a substantial answer. But in face-to-face surveys, the interviewer can be a source of item nonresponse as well. This is left for further research. I proceed with testing a few proposed strategies for item nonresponse reduction based on this review.

Over time, respondents might lose the ability to concentrate and motivation. I therefore expect more item nonresponse, the longer the interview takes. If this is true, making questionnaires only as long as necessary is a strategy to reduce item nonresponse and maintain high data quality. However, the duration of the interview might also be dependent on the respondents' ability. Overall ability of the respondent needs to be controlled for in the model.

Interferences and other people being present during the interview are expected to increase item nonresponse because they can be a source of distraction and respondents might not want to answer some questions in front of people they know. If true, interviews should be conducted without others present if possible.

Showcards are intended to make the process of answering less burdensome because respondents do not need to remember the response options. If this has an effect on item nonresponse, showcards should be provided and their use promoted to ensure a high data quality. In this operationalisation, the interviewer assesses afterwards to what extent the respondent used the showcards. Respondent ability might influence how well they understand the procedures, ability needs to be controlled for again. The interviewer assessment is likely influenced by the overall impression the interviewer has from the interview and is therefore endogenous. The overall impression of the interviewer needs to be controlled for as well.

Language difficulties are likely to inhibit the response process. For respondents that primarily speak a different language at home than the interview is conducted in, the task might be more difficult. To ensure data quality from these respondents, translating the questionnaire is an option (although costly and comes with problems of comparability). Respondents for whom this is true are likely migrant groups that have lower levels of education than the native population. Therefore, ability needs to be controlled for.

Previous research suggests that respondents might be more willing to answer to socially similar interviewers. I test this proposition for gender and age. If true and the sampling frame contains some demographic information, survey agencies could assign interviewers based on these demographics. Whether the gender of respondent and interviewer matches can be seen as random while the probability of the respondent being older or younger than the interviewer is dependent on the respondents' age, which is correlated with ability. Therefore, respondent age needs to be controlled for \footnote{Interviewer age is accounted for in the fixed effects.}.

Table \ref{table_strat} summarises the strategies to be tested. Necessary controls are respondents' ability and specifically age and the interviewers' overall assessment of the interview. Based on the theory outlined above, I do not see that inclusion of any of these control variables or the other strategies are likely to distort the effect of another variable of interest. Therefore, I am going to test all effects in a single model.

\begin{table}[h]
\caption{Reduction Strategies to be Tested}
\centering
\begin{tabular}{| l l l |}
\hline
Strategy & Mechanism & Expectation \\
\hline
Duration of the Interview & Ability, Motivation & positive \\
Interference of the Interview & Ability, Privacy & positive \\
Use of Showcards & Difficulty & negative \\
Interview not primary Language & Difficulty & positive \\
Gender Matching & Privacy & negative \\
Interviewer more than 10 years older & Privacy & positive \\
Respondent more than 10 years older & Privacy & positive \\
\hline
\end{tabular}
\label{table_strat}
\end{table}

\section{Data and Methods}

\subsection{Data}

Following this review and synthesis, I want to test the effectiveness of some potential strategies to reduce item nonresponse. I use the \textit{European Social Survey (ESS)} Round 9 collected in 2018 \citep{essericEuropeanSocialSurvey2019}. The ESS is a biannual face-to-face survey on attitudes and beliefs towards social and political topics in Europe. The data release 3.1 includes data from 49,519 respondents from 29 countries.

\subsection{Dependent Variables}

The three dependent variables are the sum of DK, the sum of refusals and the total sum of item nonresponse for every respondent. Item nonresponse is counted only in variables that are presented to all respondents, i.e. that are not affected by filtering questions. DK and refusal are not presented as response options in the ESS but are explicitly recorded as such. Interviewers are instructed to accept them without probe.

\subsection{Independent Variables}

The independent variables of interest are discussed in the previous section.

To control for ability and the interviewers' assessment of the interview, I include education (operationalised by the ISCED scale), age and squared age of the respondent as well as the interviewers' assessment of how well the respondent understood the questions, to what extent the respondent answered to the best of their ability and how often they asked for clarifications.

More information on all variables including summary statistics can be found in the \href{https://doi.org/10.17605/OSF.IO/M83GY
}{supplemental material}.

\subsection{Model}

I analyse the data using a negative binomial regression with interviewer fixed effects \citep{allisonFixedEffectsNegativeBinomial2002}\footnote{Count data models with fixed effects are quite debated \citep{wooldridgeDistributionfreeEstimationNonlinear1999, allisonFixedEffectsNegativeBinomial2002, cameronRegressionAnalysisCount2013}. For a justification of model choice, see the supplementary material.}. Standard errors are clustered by interviewer. Since the population of interest in this case are interviews and not societies, no weighting is applied. Missing data are deleted listwise.

The analyses are carried out in \textit{R} \citep{rcoreteamLanguageEnvironmentStatistical2021} using the \textit{Tidyverse} \citep{wickhamWelcomeTidyverse2019} for data handling and graphics and the \textit{fixest} package \citep{bergeEfficientEstimationMaximum2018} to estimate the regressions. The code is available  in the \href{https://doi.org/10.17605/OSF.IO/M83GY
}{supplemental material} as well.

\section{Results}

Figure \ref{fig_model} shows the incidence rate ratios (exponentiated coefficients) of the tested strategies. The exact coefficients of the tested strategies and of the control variables as well as standard errors and coefficients of model fit can be found in the table in the appendix.

\begin{figure}
\centering
\caption{Coefficients of the tested Strategies}
\includegraphics[width=\linewidth]{results_inr_ess9_figure.pdf}
\label{fig_model}
\end{figure}

The most promising ways to reduce item nonresponse seem to be boosting the use of showcards, shorter interviews and tranlating questionnaires.

With more showcard use as indicated by the interviewer (respondent used no showcards vs. used some showcards vs. used all showcards), the amount of DK reduces by about a quarter and the amount of refusal by about 18\% on average. With every additional thirty minutes interview duration, the number of DKs rises by about 21\% and the number of refusals rises by about 8\% on average. And compared to interviews conducted in the language the respondent primarily speaks at home, interviews conducted in a language different from the respondents' primary language at home show on average 42\% more DKs and 26\% more refusals.

As a general observation for all variables, the effects on the total number of item nonresponse closely mirror the effect on DK. This is not surprising since there are many more DKs than refusals. The effects on the number of refusals is typically weaker than the effect on DK but still present. This supports the idea that refusals and DKs are not perfectly separate in their meaning but not identical as well. For the variables presented so far, stronger effects on DK make substantial sense as well since they are all based on respondents' ability or difficulty of the task.

Matching respondents' gender has a small positive effect on the number of DKs. This is contrary to expectations, which suggested that matching socio-demographics of interviewers and respondents leads to a more trusting interview situation and reduces item nonresponse. The effect on refusals is not significant but should be pronounced since this strategy is partly based on the privacy mechanism. Matching by age has no significant effect. Matching respondents and interviewers therefore seems not to be a promising strategy to reduce item nonresponse.

Other people present during the interview raises the amount of refusals by 14\% in line with the reasoning that respondents do not want to answer some questions in presence of others they know. The effect on DK is not significant. Other people present might therefore influence item nonresponse more via privacy than distraction. However, due to the relatively small number of refusals, the effect on the total item nonresponse is not significant.

\section{Discussion}

I have reviewed the literature on item nonresponse and extended the cognitive \textit{satisficing} model \citep{krosnickResponseStrategiesCoping1991} with concerns about privacy to encompass all aspects that can interfere with the response process in survey interviews. Based on this new model, I have reviewed possibilities to reduce item nonresponse in face-to-face surveys.

In an empirical analysis using data from the \textit{European Social Survey Round 9}, I found that boosting the respondents' use of showcards, shortening interview durations and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse in face-to-face surveys. These strategies reduce the cognitive effort on behalf of respondents. Other people present during the interview is moderately associated with more refusals as well. Respondents are probably not willing to disclose private information in front of people they know. However, my hypothesis that respondents might trust interviewers more and share more private information if interviewer and respondent are socially similar has received no support: matching socio-demographic characteristics of interviewers and respondents seems not to be a worthwile strategy.

I want to stress though that I do not claim causal effects. Although I carefully selected control variables, I cannot rule out violations of the \textit{conditional independence assumption} which is necessary to identify causal effects with regression analysis. Most variables are influenced by respondents' ability (e.g. to understand survey procedures) as is item nonresponse. Respondents' ability is notoriously hard to measure in surveys and proxies like education and age that I have used as controls are not perfect.

A second threat to the results is the endogeneity of some variables of interest, in particular showcard use and others present during the interviews. They are measured in the interviewer questionnaire after the interview and are likely biased by the interviewers' overall assessment of the interview, including the amount of item nonresponse. I tried to control for that using other variables from the interviewer questionnaire but again, I cannot verify the conditional independence assumption to be true.

In summary, this review and analysis can only serve as a \textit{current best guess} in how to think about item nonresponse and its reduction. Future research might be able to overcome some of the shortcomings of this and other previous studies.

\section*{Acknowledgements}

I want to thank Achim Koch, Johanna Mehltretter and especially Tobias Gummer for helpful comments in multiple stages of this project. I want to thank Marcel Kappes and Thomas Gautschi for methodological discussions.

This work was supported by the University of Mannheim’s Graduate School of Economic and Social Sciences.

\singlespacing
\printbibliography

\newpage

\include{results_inr_ess9_table}

\end{document}

### 2022-07-04 inr_ESS9.R
Deleting depreciated files

@@ -1,125 +1,0 @@
-###########################################################################
-### Item Nonresponse in ESS Round 9
-### Author: Malte Grönemann
-### Version: R 4.0.5 on Linux
-###########################################################################
-
-library(fixest)
-library(tidyverse)
-library(haven)
-
-
-# import of the raw data and joining
-# necessary to change encoding on Linux/OSX, remove if on Windows. 
-# See https://haven.tidyverse.org/reference/read_dta.html#character-encoding
-ESS9 <- full_join(read_dta("ESS9e03_1.dta", encoding = "latin1"), 
-                  read_dta("ESS9INTe03.dta", encoding = "latin1"), 
-                  by = c("idno", "cntry"))
-
-# calculating the numbers of refusals and DK by interview
-# See https://haven.tidyverse.org/articles/semantics.html#tagged-missing-values-1
-# ESS coding: Not applicable -> a, Refusal -> b Don't know -> c, Not available -> d
-count_refusals <- ESS9 %>% 
-  select(nwspol:impfun) %>% # only substantial variables
-  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>% # excluding vars affected by routing
-  mutate(across(.fns = ~ is_tagged_na(.x, tag = "b"))) %>%
-  rowwise() %>%
-  transmute(n_refusals = sum(c_across()))
-
-count_dk <- ESS9 %>% 
-  select(nwspol:impfun) %>%
-  select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE)) %>%
-  mutate(across(.fns = ~ is_tagged_na(.x, tag = "c"))) %>%
-  rowwise() %>%
-  transmute(n_dk = sum(c_across()))
-
-# total numbers of DK and REF
-sum(count_dk$n_dk)
-sum(count_refusals$n_refusals)
-
-# selecting and if necessary transforming variables for analysis
-analysis <- ESS9 %>%
-  transmute(`Age (in 10 years)` = agea / 10, # respondent
-            `Female` = gndr - 1,
-            `Education (ISCED)` = ifelse(eisced < 10, eisced, NA),
-            `Ethnic Minority Member` = blgetmg * -1 + 2,
-            `Member of discriminated Group` = dscrgrp * -1 + 2,
-            `Political Interest` = (polintr-1) * -4/3 + 5,
-            `People have Political Influence` = psppipla,
-            `Health` = health * -1 + 6,
-            `Generalised Trust` = ppltrst/2,
[…]

### 2024-01-31 Reviews MDA Round 1/response_reviews.tex
revised text after second round of reviews

\documentclass[a4paper, 12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to the Reviews: “How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European
Social Survey”}

\begin{document}

\maketitle

Dear editors and reviewers of MDA,

Thank you for your reviews and I am very grateful for the opportunity to improve upon my work. And I am sorry for the long time it took me to address your criticisms. I have organized my response in sections in roughly chronological order of the article. I hope I have addressed your criticisms to your satisfaction. When I disagree with a comment of the reviewers or am not able to incorporate a requested change, I provide my reasons here as well.

\section{Language and Structure}

\begin{quotation}
``All reviewers commented on the language and one very helpful reviewer even added an annotated file with suggestions for language changes in your rewrite. We have added this file. Also, it was suggested to ask a native speaker to go over the final manuscript. Several reviewers commented that the manuscript was unnecessarily complicated and long and suggested that you restructure it. Finally, it was suggested to incorporate the important Tables directly in the manuscript and not in a supplement. When rewriting the manuscript, we ask you to focus on this and take the detailed comments of the reviewers into account.'' (Editor letter, p. 1)
\end{quotation}

I have thoroughly rewritten all sections of the article and checked grammar and spelling throughout. I hope that the improvements in language are to your satisfaction.

Regarding length, I am not sure where to cut it down. And the article is still within the word limits of a research note. As large parts of the article are literature reviews, it indeed might feel slow but that is inherent to literature reviews. If you still consider the manuscript too long, I would ask you to state what can be omitted and why.

Similarly, the general structure going from theory to strategies to empirical analysis is stringent to me so I do not know what you expect me to change here. If you still consider changes in structure necessary, please give me more details. However, I restructured the section on the tested strategies. This section only presents the selected strategies and why I expect them to reduce INR. Operationalization and the selection of controls have been moved to the data and methods section. And some reviewers have requested to connect the different parts more, especially when switching from review to analysis. I tried to smooth out the transitions between sections and make the article more coherent.

The regression table is now placed within the text, the exact design would be part of the layout though.

\section{Conceptualisation of INR}

In my article, I conceptualized four constructs related to the probability of item nonresponse, cognitive ability, task difficulty, motivation, and privacy concerns. Reviewer D writes:

\begin{quotation}
``I do not agree with this conceptualization, as it makes identifying the source difficult. Rather, it would be helpful to distinguish sources of item NR overall or within each construct the author proposes. For example, one can conceptualize respondents, questions, questionnaires, interview protocols/settings, and interviewers as sources of the item NR [...]. For example, motivation can be a product of all sources. Lumping these sources into the constructs is likely to make coming up with strategies for reducing item NR extremely challenging.''
\end{quotation}

I disagree with this notion. Providing a nonresponse to an item is always an action of the respondent. And a specific respondent (with its characteristics) is asked to do a task in a specific situation. This task might be more or less difficult or boring etc. for this respondent in this situation. Survey design can influence the task and situation, but not directly item nonresponse, only indirectly. I think that it is necessary to conceptualize item nonresponse in terms of the respondent. Questionnaire design etc. are not ``sources'' of INR but influence the sources. They are not acting entities and placing ontological meaning onto them misrepresents the survey response process.

Reviewer D again writes in a similar vein: ``Privacy concern is hypothesized to come in in the editing process, but I argue it may also come during the comprehension stage.''. Again, I have to disagree with this notion. Concerns about one's privacy are not a matter of comprehension. They are right that such a concern can become salient to the respondent before coming up with an answer. But Tourangeau et al. (2000) explicitly allow that although the steps of responding are typically in that order, they do not need to be. Editing may therefore kick in right after comprehension.

Additionally, this conception is based on the conceptions of Krosnick (1991) and Tourangeau, Rips, and Rasinski (2000) which are widely cited in this literature. Such objections are therefore going far beyond what this paper aims for and is capable of as they would require a different conceptualization of survey response entirely.

Since only one reviewer had such fundamental objections and the editors decided to allow me a revision even though such a fundamental part of the paper is objected to, I will not change this conceptualization. I have tried to make the position that the respondent is the acting entity clearer in section 2 though.

\section{Extension of the Satisficing Model}

Reviewer D claims that I fail to deliver the extension of the satisficing model promised in the abstract. This is somewhat contrasted by Reviewer E though: ``Your novelty is the introduction of ‘privacy concerns’ in the model developed by Krosnick, so you should give more visibility to it. Therefore, my first suggestion is to make more evident the novelty you are bringing to the field.''

So, Reviewer E thinks that I do provide a meaningful extension of Krosnick's satisficing model. Maybe the reason Reviewer D believes that I did not is indeed due to the lacking visibility.

My aim with this extension (and the formalization) is to synthesize and summarize results from the literature on INR. I believe that without synthesis and theoretical models, we can only achieve a piece-meal understanding of such processes and result in ad-hoc explanations or survey designs.

This holds for the formula as well. It is not intended to be an actual formula providing a point estimate of the probability of item nonresponse given some measured inputs (although it would of course be better). In terms of the aim of the formula, I again follow Krosnick. It serves as a summary of what determines INR and how these concepts interrelate. The inputs, therefore, need not be measurable on a proper scale (as commented on by Reviewers D and F). I made this fact explicit in the article and changed the formula so that it reads ``the probability of harmful item nonresponse is a function of...'' instead of claiming to have a full formula of the probability.

Reviewer D also asked why I chose a max function in the formula. The reason is that I conceptualize privacy concerns as being independent of the other concepts. The probability will then either be driven by the relations of task difficulty, cognitive ability, and motivation OR privacy concerns depending on which of those exceeds the other. Reviewer E also requested to better explain the aim and meaning of the formula so I tried to make it clearer in section 2.

Reviewers D and F also point out that the formula is not picked up in the empirical analysis. This is a result of a shift from literature review and theoretical synthesis first to practical applications in survey design following from that model later. Reviewer F mentions that the link between those two parts is indeed weak. I tried to point out the link between the theoretical model and empirical analysis accordingly (introduction to section 4 and section 5.3).

\section{Literature Review}

\begin{quotation}
Lines 193-195: by ‘changes in response categories’ do you mean ‘changes in response scales’? Routing and filtering are associated with higher non-response in PAPI, using CAPI mode may reduce routing and filtering problems to zero. (Reviewer E)
\end{quotation}

I indeed mean changes in response scales. I changed the wording. And I added that this result is likely different between modes.

\begin{quotation}
`` Regarding ability, there is much more to say. The environment of the interview is important, but ability can (and must) be assessed before going into the field by a pilot study, in order to test the comprehension of the questions, the rates of item non-response, the impact of the order of the questions (contamination), among other aspect. (Reviewer E)
\end{quotation}

I would categorize your comments not under ability but task difficulty as they relate to survey design and not respondent characteristics or the interview situation. But you are right that review rounds and pilots are important tools to ensure survey quality also in this regard and I added a paragraph on them in section 3.1.

\begin{quotation}
“Matching education has no effect (Vercruyssen, Wuyts and Loosveldt, 2017;…” not quite…what the authors say is that “the effects of education level (mis)matching could unfortunately not be tested.” What had no effect was the education level of the interviewer.'' (Reviewer E)
\end{quotation}

Thank you for the correction, I removed the citation (Vercruyssen, Wuyts and Loosveldt, 2017) but Silber et al. do test it and find no effect.

\section{Don't Know as a substantive Answer}

\begin{quotation}
``the author acknowledges that 'don't know answers' for some respondents actually be the most applicable answers (i.e. "because they do not know about the content of the question or are unable to remember an event"), initially this is not mentioned, and also when discussing the four constructs related to the probability of item nonresponse, this issue is ignored. I think the author could put more emphasis on the difference between satisficing and privacy-originated sources of item nonresponse on the one hand, and substantive don't know answers on the other. [...] I do not agree with the author's argumentation that in cases where DKs are genuine, this is "a product of the respondents' ability", since especially in attitude items, it may often happen that before the survey respondents have never thought about the issue being questioned.'' (Reviewer A)
\end{quotation}

Reviewer A makes a good point here. The reason why I do not consider DK as a genuine answer is simply because genuine answers are unproblematic. This paper aims to improve survey quality and therefore seeks to reduce \textit{problematic} INR. However, I have clarified why I do not consider genuine DKs in detail and removed the argument that they are due to respondents' ability. Accordingly, the definition of item-nonresponse as ingenuine answers has been stressed.

For the analysis, I do not expect that the presence of genuine DKs in the dependent variable would bias some of the estimates. At least, no such mechanism came to my mind. I, therefore, did not change the analysis based on this comment. Genuine DKs only increase variation in the dependent variables but do not bias the results, as mentioned in section 5.2.

\section{Data and Variables}

Reviewers C and D asked for the data to be introduced more thoroughly and especially to add information on protocols and contexts relevant to INR. I added some fundamental information about the ESS in section 5.1 but I still wanted to keep it short. I believe the ESS is a very widely known data set and I have seen in other research notes in MDA that data descriptions are short even with lesser-known data. I added a reference for further information though and more detailed information on the data has been added to the supplementary material.

The latter point of Reviewer D to provide more information on protocols and procedures relevant to the present study is very important though. I have added information about the policies regarding the use of showcards, the intended length of the interview, and interviewer behavior in dealing with respondents and especially item nonresponse in sections 5.1 to 5.3.

Reviewer D also asked why I restrict respondent characteristics to education and age. Since my analysis shifts from the theoretical model to strategies for reducing INR on the survey level, I only need to include respondent characteristics if they threaten to confound my analysis of strategies. I use both of them as proxies for cognitive ability but I do not need further respondent characteristics as controls. In section 5.3, I have stressed this point again. No other reviewer complained about the selection of controls so I have not changed them.

Reviewer E pointed out that matching respondent and interviewer characteristics in terms of ethnic/national origin could be important. That is true and I added this to the literature review of strategies. But since ethnic/national origin is not measured of interviewers, I cannot assess its effect in my analysis, unfortunately.

\section{Statistical Analysis}

\begin{quotation}
``it is important to note that attitude items often have higher levels of item non-response than factual items on the same topic, reflecting opinion censoring, which is actually higher among higher educated respondents than lower educated respondents. Therefore, I suggest that the type of item, attitudinal or factual, is also taken into account in the analysis of the ESS data.'' (Reviewer A)
\end{quotation}

Reviewer A makes an important point here and I have added this to the review. However, the empirical analysis aims to test strategies to reduce item nonresponse at the survey level, not to understand influences on the item level. It might indeed be the case that these strategies would have a different effect depending on the type of question (attitudinal or factual). But to disentangle this, I would need to conduct separate analyses on different subsets of the items. Given that multiple reviewers already called the manuscript too long, I would not extend the analysis. This is an important shortcoming of the paper though and I added this to the discussion. If strategies differ in their effects depending on the type of question, then the external validity beyond the mix of question types as in the ESS could be limited.

\begin{quotation}
Unless reading the supplemental materials, readers will not know why negative binomial model is used. Also, why interviewer effects are analyzed as fixed? “Standard errors are clustered by interviewer.” on p.15 does not make sense. (Reviewer D)
\end{quotation}

I added a short reason for the chosen model in section 5.4. I use interviewer fixed effects because I want to get rid of variation in levels of INR between interviewers since this part of the variation is not relevant to the tested strategies. Standard errors are typically clustered by the variable denoting the fixed effects to prevent heteroscedasticity within the clusters. As now pointed out in the supplementary material, this is common practice and even the default in the R package used. I compared the clustered standard errors to bootstrapped standard errors and found no substantive differences. I extended the discussion on the fixed effects and standard errors in the supplementary material and highlighted the supplementary material in the context of model selection as a compromise to the comment of Reviewer C to discuss model selection in the text and the need to keep the article brief.

Reviewer D asked about missing data in my analysis pointing out that I have a large amount of missing data. As mentioned in the supplementary material, there are two major sources of missing data in my analysis. The first one is the age of the interviewer is often missing and Romania used the wrong format for the interviewer's age altogether. I resultingly had to completely exclude Romania. But this would only be consequential if the nonresponse in the age of the interviewer is related to the interaction between respondent and interviewer since I control for interviewer effects using the fixed effects. The second source of reduced observations is a lack of variation within interviewers in the fixed effects regression. Since a within-estimator needs within-variance, interviewers without within-variance are excluded from the analysis. This is particularly often the case for the number of refusals since the number of refusals is relatively low. Although such a large loss of observations is indeed unfortunate, they can be assumed to be random and therefore not bias the results. The levels of item nonresponse in the variables used for analysis is always reported with descriptive statistics in the supplement and are not particularly high. The largest part of the lost observations can therefore be assumed to be missing at random and does not bias the results. This discussion is extended in the supplementary material as well.

\section{Length of Interview as a Covariate}

\begin{quotation}
 ``Since length of the interview may vary depending on the applicability of questions, I am not sure what the mechanism here is: respondents for whom more questions are applicable will have longer duration and are different from those who have fewer applicable questions due to filter questions [...]. Since interview length cannot be disentangled from respondent characteristics, it is completely unclear what is causing the positive effects of interview length on item nonresponse.'' (Reviewer A)
\end{quotation}

As discussed in Section 4, there are two reasons for longer interviews. The first reason is that the respondent takes longer to answer which is a result of the respondent's cognitive ability. The second reason is that a higher number of questions asked will lead to the respondent losing interest and getting tired. The strategy to be tested is that a shorter questionnaire prevents INR via the second mechanism. But to be able to assess this, I need to control for cognitive ability.

However, based on this comment by Reviewer A, I have realized that I have a better measure for questionnaire length than time duration, which is the number of items not applicable to the respondent. I have therefore exchanged the duration with the number of items. This variable should have a less direct connection to cognitive ability. The comments on interview duration by Reviewer E resultantly do not apply anymore.

The quoted comment by Reviewer A claims that the people with a different number of items might be substantially different from those with many items not applicable. But to bias the results, they would need to differ concerning motivation, cognitive ability, or privacy concerns following the presented theoretical model. I was not able to think of a mechanism for why that would be the case in general, therefore it would depend on the specific questions that would have been asked. I checked descriptively whether there is some form of a relation between the number of applicable items and education (as a proxy for ability) and found no strong bivariate relation (see the supplementary material). Unfortunately, I do not have proxies for motivation and privacy concerns to explicitly check them but I think that a spurious correlation between ability and the number of applicable items would have been the most likely case. I would therefore argue that the number of applicable items is a reasonable test of the mechanism but caution is of course advised.

However, the results indicate that the number of applicable items has a negative effect on INR and this result is robust across model specifications that also include duration. I, therefore, need to retract my claim from the previous version that longer questionnaires increase the number of INR. Surprisingly, I find a negative association between INR and longer questionnaires and I do not know how to interpret this result myself. I added a short disclaimer to the discussion section. One might argue though that it is in fact the time they are required to concentrate that is more important than the number of questions. But this would again run into the problems outlined above. If you have any recommendations or ideas on this issue, please let me know.

\section{Miscellaneous}

\begin{quotation}
 Be careful with contradictions. For instance, in lines 49-51 (and 368-3689) you write: “Matching respondents’ and interviewers’ gender and age seem not to be efficient tools to reduce item nonresponse.”; but in lines 234-236 it is said that “Vercruyssen, Wuyts and Loosveldt (2017) find less item nonresponse when matching age of interviewers and respondents and matching gender reduces item nonresponse for males and increases it for females”; and in lines 276-277, “Previous research suggests that respondents might be more willing to answer to socially similar interviewers.” (Reviewer E)
\end{quotation}

Those are not contradictions but differing results from different studies. Some scholars have hypothesized this link and could show it empirically but in my analysis, this did not seem to matter. The conflicting statements report an inconclusive state of research.

\begin{quotation}
``The motivation for this particular study is not clearly provided in Introduction. Rather, Introduction mixes in the findings from the study.'' (Reviewer D)
\end{quotation}

I tried to stress the contribution in the Introduction. However, summarising the general findings in the introduction is quite common in academic papers. If requested, this can be removed but I do not see the general issue with that practice.

\begin{quotation}
``The discussion section would gain if the novelty of this study was stressed
Finally: the tendency of many surveys (due to high costs and low response rates) is to change from face-to-
face to self-administrated modes (PAPI and CAWI). Consider discussing a bit this incoming shift and how your
final recommendations could be adapted.'' (Reviewer E)
\end{quotation}

Thank you for these suggestions, I have tried to satisfy your requests.

\end{document}

### 2024-01-31 Reviews MDA Round 2/response_reviews2.tex
revised text after second round of reviews

\documentclass[a4paper, 12pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{Response to the Reviews (Round 2): “How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European
Social Survey”}

\begin{document}

\maketitle

Dear editors and reviewers of MDA,

Thank you for your reviews. I am very happy that the improvements are acknowledged, I myself think that the paper massively improved in the first round of reviews. I therefore get back to your minor requests as soon as possible. I again structured my response into sections.

For clarity, I will refer to the reviewer who provided the annotated PDF as reviewer 1 (R1), the other is denoted reviewer 2 (R2) respectively. As the reviews are based on the last version, page numbers will usually be from the old document ("OD"). If I specifically refer to the revised document, I add "RD".

\section{Language}

Both reviewers rightfully pointed to some ambiguous and lengthy sentences. R2 generally found parts throughout the text confusing. I therefore had a read myself again and improved language throughout. I will report only on major changes and the reviewer comments here though. Most of the time, I will just present the updates made in the RD.

\begin{itemize}
\item R2 1a: "I will therefore not distinguish between refusal and DK"
\item R2 1b: added "than closed single choice items"
\item R2 1c and R1 p.9: "How question design affects levels of item nonresponse is very well understood and differences between questions constitute the largest part of the variance in item nonresponse (Olson, Smyth and Ganshert, 2019)."
\item R2 1d: "To reduce task difficulty for members of language minorities, the questionnaire can be translated so that respondents can take the interview in the language they are most proficient in." The source is given in the following sentence: Behr and Shishido (2016)
\item R1 p.12: it is indeed a personal though: "I could imagine that some of them may be mistrustful towards interviewers due to racist experiences and a fear of discrimination."
\item R1 p. 16: "Based on the theory outlined above, I do not expect that the inclusion of any of these control variables or the other strategies is likely to distort the effect of another variable of interest."
\item R1  p. 26: R1 pointed out that homophily is incorrect as a term here as it refers to observed network patterns, not preferences. I changed it to "strategies based on the idea that respondents trust socially similar interviewers".
\end{itemize}

\section{Nonresponse and Editing}

\begin{quotation}
R2 Comment 2: My previous comments about item nonresponse with the editing stage has not been address. The paper still asserts, “When editing an answer, respondents may have concerns about their privacy.” on p.6. Respondent may have privacy concerns when hearing the questions, far before reaching the editing stage.
\end{quotation}

I think that this is a simple misunderstanding. As mentioned on p.4, respondents can jump between stages in the survey response process. The stages do not need to be in order, although they often are. In such a case, it is indeed likely that respondents jump from comprehension to editing. I added two sentences to the paragraph to make this clear:

\begin{quotation}
Respondents will likely have such privacy concerns immediately when they hear a sensitive question and jump from the comprehension stage to the editing stage in the survey response process. They probably refuse to answer before an honest answer has been formed.
\end{quotation}

\section{Formula}

\begin{quotation}
R2 Comment 3: I don’t find the formula for $P_{INR}$ necessary for introducing the four concepts that are the core of this paper. It is unfortunately distracting. Further, without a quantification of difficulty/(ability*motivation) and privacy, you certainly cannot solve the max function involving these two terms.
\end{quotation}

The associate editor highlighted these concerns as follows:

\begin{quotation}
"Comment 3 about the formula needs some attention, in the answers accompanying the revised manuscript, the author states, that the formula is not a n actual statistical formula but a help in summarizing and defining interrelationships between concept. This should be very clearly stated in the text!  And perhaps the word formula should be omitted. Perhaps something like ‘the interrelationship between the concepts can be formalized as…’ or something similar in the own words of author.  Now it still looks like an actual statistical formula."
\end{quotation}

I disagree with R2 that the formula is unnecessary and distracting. If I would just state the concepts as such, their nonlinear effects and interdependence cannot be conveyed. As this is the primary function of the formula, and not to actually provide a way to predict nonresponse, the concepts need no proper way of measurement as well.

I hope that in the revised version, this intention is now coming across even more clearly. As recommended by the associate editor, I exchanged the word "formula" for "theoretical model" to avoid confusion with a statistical formula and I introduce the equation as a formalisation and summary. The last paragraph of that section also highlights that the equation is not intended to give predictions and that the concepts need not to be measured for the purpose of it.

\section{Aim and Causality}

\begin{quotation}
R2 Comment 6: The authors mention that they will not speaking causal effect on p. 17. Then, what is this paper for? The paper proposes the four concepts as “sources” of item nonresponse, which implicitly suggests a causal mechanism.
\end{quotation}

The empirical part of the paper does not set out to test whether the four concepts have a causal relationship to item nonresponse. The literature review shows overwhelming empirical evidence that they do. My empirical contribution aims to derive possible strategies to reduce item nonresponse and test whether they indeed are associated with a change in item nonresponse. So, R2 is right that the theoretical model suggests causal mechanisms for these strategies. That is the whole reason why I believe these strategies are worth investigating.

However, my empirical analysis needs to fulfill the conditional independence assumption to estimate an unbiased point estimate for the causal effect of a strategy, which is a very strong assumption. I expanded the last paragraph of the section on controls, where I make the claim R2 mentions, to better explain why I do not call my estimates causal but still believe them to be informative. I also made an effort to communicate my aim of the empirical analysis more clearly in the first paragraph of the section "Strategies to be tested".

\section{Standard Errors}

\begin{quotation}
Comment 7 (R2): "“Standard errors are clustered by the interviewer” on p. 18. Standard errors cannot be clustered. Please clarify. "
\end{quotation}

I agree with you that the name "clustered standard errors" is probably not the best, standard errors cannot be literally clustered. However, I did not coin the term and it is widely used in econometrics by now. I also provided a specific reference for this recommendation and I have outlined the reason why the standard errors are "clustered" by interviewer in the supplementary material (see p. 10) before submitting the OD.

As it is an established term and quite well known in econometrics, I am hesitant to changing the wording, as the more appropriate name "Liang-Zeger standard errors" is much less common.

\section{Table 1}

\begin{quotation}
R2 Comment 4: “Expectation” column in Table 1. Expectation for what? Expectation in nonresponse or nonresponse reduction?
\end{quotation}

By expectation I refer to the expected sign of the coefficient if the strategy works. R1 also requested an explanation of table 1 in the text. I added a paragraph before the table that explains it:

\begin{quotation}
Table 1 summarises the selected strategies that I am going to test in my empirical analysis. The second column shows which concepts play a role in the hypothesized mechanism linking the respective strategy to item nonresponse. The third column gives the expected direction of the relationship between strategy and item nonresponse, e.g. the longer the questionnaire, the more item nonresponse. These are also the expected signs of the coefficients if the strategies work as imagined.
\end{quotation}

\section{The two meanings of DK}

\begin{quotation}
R1 p.15: Also in your to the reviewers you state that they do not bias the results. In my opinion you should explain exactly why this is the case...
Assuming they are evenly distributed and uncorrelated to any of the independent variables or covariates?
\end{quotation}

R1 is right, I assume the two mechanisms that can lead to DK, geneuine and harmful, to be uncorrelated so that my estimates are unaffected. I cannot think of a way how the possibility of genuine DK answers should bias my estimates. I have extended the respective paragraph in the subsection on dependent variables to make clear that this is an assumption on my behalf:

\begin{quotation}
Secondly, an additional mechanism that generates item nonresponse may increase the overall variation in the dependent variables but if it is uncorrelated to the other mechanisms, no bias in estimates is to be expected. I do not know how the possibility of genuine DK answers could interfere with the other mechanisms. I therefore assume that they are uncorrelated.
\end{quotation}

\section{Controls}

\begin{quotation}
R2 Comment 5: “Interferences and other people being present during the interview as well as whether respondent and interviewer have matching gender can be regarded as unrelated to any of the four theoretical concepts outlined in the model.” on p.16 Why?
\end{quotation}

I am simply not aware of a mechanism potentially biasing these estimates. I changed the wording to better reflect this reasoning:

\begin{quotation}
I am not aware of any mechanisms that could lead to biased estimates for the effects of interferences and other people being present during the interview as well as whether respondent and interviewer have matching gender.
\end{quotation}

\end{document}

### 2024-01-31 inr_ESS9_supp.Rmd
revised text after second round of reviews

@@
 title: "Supplement to 'How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey'"
-author: "Malte Grönemann, University of Mannheim"
-date: "04/07/2022"
 output: pdf_document
@@
 
+This supplementary material provides the interested reader with more detail on the data, data manipulation, descriptive statistics, and choices made in the analysis.
+This serves to keep the original text short and still provide full transparency on data, code, and analysis.
+
 # Version and Packages
@@
 library(tidyverse) # 1.3.1
-library(haven) # 2.5.0
-library(qqplotr) # 0.0.5
+library(haven) # 2.5.2
+library(qqplotr) # 0.0.6
 library(fixest) # 0.10.4
+library(kableExtra) # 1.3.4
 ```
 
-# Data Manipulation
+# Data
+
+Following the literature review and synthesis in the article, I want to test the effectiveness of some potential strategies to reduce item nonresponse.
+I use the \textit{European Social Survey (ESS)} Round 9 collected between August 2018 and January 2020.
+The ESS is a biannual face-to-face trend survey on attitudes and beliefs towards social and political topics in Europe established in 2001.
+In each country and round, the ESS draws a new random sample of the residential population of 15 years and older aiming for a minimum response rate of 70\%.
+Most countries use computer-assisted personal interviews for data collection and the questionnaire is designed to take about one hour.
+The data release 3.1 includes data from 49,519 respondents from 29 countries (Austria, Belgium, Bulgaria, Croatia, Cyprus, Czechia, Denmark, Estonia, Finland, France, Germany, Hungary, Iceland, Ireland, Italy, Latvia, Lithuania, Montenegro, Netherlands, Norway, Poland, Portugal, Serbia, Slovakia, Slovenia, Spain, Sweden, Switzerland, and the United Kingdom).
+For more information on the data, visit \href{https://www.europeansocialsurvey.org}{europeansocialsurvey.org}.
+I will provide specific information about ESS policies and design in this supplement when needed for the variables discussed.
+
+## Data Manipulation
 
 `ESS9e03_1.dta` is the main data file from the ESS 9 version 3.1 and `ESS9INTe03.dta` is the interviewer questionnaire.
+The former contains predominantly the answers of the respondents but also weights, automatic time stamps, and other metadata.
+The latter contains information about the interviewer and the answers of the interviewer to a short post-interview questionnaire.
 The data are available after registration here: https://www.europeansocialsurvey.org/data/download.html?r=9
-*idno* is a respondent identifier and *cntry* is the country abbreviation.
 
-It is necessary to change the encoding on Linux/Mac to *latin1*, remove the option if on Windows.
-See https://haven.tidyverse.org/reference/read_dta.html#character-encoding
+I combine the two data sets into a single one by matching the ID of the respondent (*idno*) and the country abbreviation (*cntry*).
+
+It is necessary to change the encoding on Linux/Mac to *latin1* to import the data correctly, remove the option if on Windows (see https://haven.tidyverse.org/reference/read_dta.html#character-encoding).
 
@@
 
-I use three dependent variables: the number of refusals by a respondent, the number of Don't know (DK) and their sum.
-I sum them over all questions applicable to all respondents.
+I use three dependent variables: the number of refusals by a respondent, the number of Don't Know (DK), and their sum.
+I take the sum of all questions applicable to all respondents.
 That means, all questions only asked a subset of respondents due to previous answers or survey experiments are left out.
 
-As background information, refusal and DK are not read out to the respondents in the ESS or on the showcards.
+As background information, refusal and DK are not read out to the respondents in the ESS and are not shown on the showcards.
 But interviewers have them as distinct options within their CAPI instrument.
-Interviewers are advised to accept them without probe.
-Whether the respondent refuses or says DK is therefore an interpretation of the interviewer.
+Interviewers are advised to accept them without probing.
+Whether the respondent refuses or says DK is therefore to some extent based on the interpretation of the interviewer.
 
-`select(nwspol:impfun)` selects only the variables that are based on questions asked the respondent and gets rid of the interviewer questionnaire, weights etc. for the calculation of item nonresponse.
+The command `select(nwspol:impfun)` selects only the variables that are based on questions presented to the respondent and gets rid of the interviewer questionnaire, weights, etc.
+for the calculation of item nonresponse.
 
@@
 
-- .a - not applicable
+- .a - not applicable (due to filtering, missing by design)
 - .b - respondent refused
@@
 
-To have a comparable set of items for all respondents, I only select the ones applicable to all: `select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE))`
+To have a comparable set of items for all respondents, I only select the ones applicable to all respondents and exclude items that are affected by filtering questions: `select(where(~ any(is_tagged_na(.x, tag = "a")) == FALSE))`
 
@@
 
-The standard deviation of DK is `r sd(count_dk$n_dk)` and the of refusals is `r sd(count_refusals$n_refusals)`.
-Although for the regression the variance of the residuals matters, this is a first sign of overdispersion.
+The standard deviation of DK is `r sd(count_dk$n_dk)` and the standard deviation of refusals is `r sd(count_refusals$n_refusals)`.
+Although ultimately the variance of the residuals matters for the regression, this is a first sign of overdispersion.
 
@@
 
+## Number of Applicable Items
+
+I hypothesize that longer questionnaires generate proportionally more item nonresponse because respondents' motivation decreases over time and respondents might get tired/exhausted over time reducing their cognitive ability.
+To test this, I generate a variable that measures how many items apply to the respondent based on the missing code .a.
+The more items that are not applicable, the fewer questions the respondent is asked.
+
+For better interpretation, I do not use the number of not-applicable items but the number of applicable items.
+To get this, I subtract the number of not-applicable items from the number of total items.
+The number of total items is `r ESS9 %>% select(nwspol:impfun) %>% length()`.
+
+```{r na_items}
+count_applicable <- ESS9 %>% 
+  select(nwspol:impfun) %>%
+  mutate(across(.fns = ~ is_tagged_na(.x, tag = "a"))) %>%
+  rowwise() %>%
+  transmute(n_applicable = 545 - sum(c_across()))
+```
+
 ## Preparation for Analysis
 
-The variables from the interviewer questionnaire still contain missing values as number codes 8 and 9.
+The variables from the interviewer questionnaire still contain missing values coded as 8 and 9.
 They are set to NA if their value is above 6.
-Bulgaria used the wrong format for interviewer age and therefore adds many missings to it and is resultantly excluded.
+Bulgaria used the wrong format for the interviewer's age and therefore adds many missings to it and is resultantly excluded.
 The *other* category of ISCED (number code 55) is coded to NA as well.
-The length of the interviews are top coded to 3 hours because there were errors with the CAPI instruments time stamps in some occasions (see ESS 9 codebook).
+The length of the interviews is top coded to 3 hours because there were errors with the CAPI instrument time stamps on some occasions (see ESS 9 codebook).
+
+Out of an abundance of caution, I merged the interviewer ID with the country abbreviation to make sure that there are no wrong fixed effects if two interviewers have the same ID in different countries although this should not exist.
 
@@
 analysis <- ESS9 %>%
-  zap_label() %>% zap_missing() %>% zap_formats() %>% # getting rid of Stata formatting
+  zap_label() %>% 
+  zap_missing() %>% 
+  zap_formats() %>% # getting rid of Stata formatting
   transmute(age = agea, # respondent
@@
             duration = ifelse(inwtm < 180, inwtm / 30, 6),
+            n_applicable = count_applicable$n_applicable / 10,
             interference = ifelse(preintf < 6, preintf * -1 + 2, NA),
@@
             bestab = "Answered to best Ability",
+            n_applicable = "Number of applicable Items (10 Items)",
             duration = "Duration of Interview (30 Min)",
@@
 ```
+
 ## Descriptive Statistics
@@
 
-Except of the use of showcards and the duration of the interview, all variables of interest are dummys:
+Except of the use of showcards and the number of applicable items, all variables of interest are dummys:
 
@@
 
-`r mean(analysis$notprimlang, na.rm = T)*100` percent of the interviews were conducted in a language different from the one the respondent primarily speaks at home. There are no missings in this item.
+`r mean(analysis$notprimlang, na.rm = T)*100` percent of the interviews were conducted in a language different from the one the respondent primarily speaks at home.
+There are no missing data in this item.
 
@@
 
-The duration of the interview is automatically computed by the CAPI instrument based on start and end time stamps.
-There have been some issues with the time stamps in some countries though, resulting in some unusually long durations.
-I top-coded them to 3 hours and rescaled the variables to 30 minutes.
+The number of applicable items is described below.
+A potential bias in the check of the mechanism that longer questionnaires increase INR because respondents get tired would be if the respondents that are asked more or less questions differ systematically in their ability as this could confound the expected decrease in motivation.
+The scatter plot of the number of applicable items and education as a proxy for ability serves to check that.
+It does not seem to be the case that a different number of applicable questions is related to ability and differences in other respondent characteristics should not affect the outcome according to the theoretical model outlined in the article.
 
-```{r duration}
-summary(analysis$duration)
+```{r n_applic_descr}
+summary(analysis$n_applicable)
 
 ggplot(analysis) +
-        aes(x = duration) +
+        aes(x = n_applicable) +
         geom_histogram() +
-        labs(x = "Interview duration", y = "")
+        labs(x = "Number of applicable Items (10 Items)", y = "")
+
+ggplot(analysis) + 
+        aes(x = educ, 
+            y = n_applicable) + 
+        geom_jitter() +
+        geom_smooth(method = "loess", 
+                    se = FALSE) +
+        labs(x = "ISCED",
+             y = "Number of applicable Items (10 Items)")
 ```
@@
 
-`r table(analysis$educ, useNA = "always")`
+The following table shows absolute frequencies of the educational levels:
 
-All control variables from the interviewer questionnaire are measured on a five point scale: Never, Almost never, Now and then, Often, Very often with number codes from 1 to 5.
+`r kable(table(analysis$educ, useNA = "always"))`
+
+All control variables from the interviewer questionnaire are measured on a five point scale: Never, Almost never, Now and then, Often, Very often. The respective number codes range from 1 to 5.
 
@@
 
-I analyse the data using a negative binomial regression with interviewer fixed effects .
-Standard errors are clustered by interviewer.
-Since the population of interest in this case are interviews and not countries, no weighting is applied.
+I analyze the data using a negative binomial regression with interviewer fixed effects.
+I choose this model because my dependent variable is a count variable that has the typical skewed shape of count variables making it unfit for linear regression.
+I chose a negative binomial link function instead of a Poisson regression because there are signs of overdispersion and because it is generally recommended.
+A more thorough discussion on model choice can be found below.
+
+I use interviewer fixed effects to control for any variation on the interviewer level.
+Standard errors are clustered by the interviewer as well.
+This is common practice with fixed effects because the fixed effect only gets rid of the differences in means between groups but there might be differences in variance between groups as well.
+Not acknowledging potential differences in variance between groups in fixed-effects regression would lead to heteroscedasticity and therefore wrong statistical inference.
+Normal parametric standard errors tend to be too large (Cameron and Trivedi 2013).
+Clustered standard errors are therefore the default in the fixest package (see https://cran.r-project.org/web/packages/fixest/vignettes/standard_errors.html).
+I replicated the regressions in Stata using bootstrap standard errors (not shown) and only get marginally different results, likely from differences in the optimization algorithm.
+
+The ESS does not draw simple random samples in all countries but the country samples are often stratified.
+In usual analysis, this requires the use of design weights acknowledging different probabilities of being in the sample.
+It is additionally recommended to use weights to correct for different sampling probabilities due to unit nonresponse.
+But since this analysis is not concerned with making inferences about the population of countries but inferences about realized interviews, no weighting is applied.
 
 The missings are deleted listwise.
+There are two major sources of missing data in my analysis.
+The first one is the age of the interviewer is often missing and Romania used the wrong format for the interviewer's age altogether.
+I resultingly had to completely exclude Romania.
+But this would only be consequential if the nonresponse in the age of the interviewer is related to the interaction between respondent and interviewer since I control for interviewer effects using the fixed effects.
+Although such a large loss of observations is indeed unfortunate, they can be assumed to be random and therefore not bias the results.
+The levels of item nonresponse in the other variables used for analysis is always reported with the descriptive statistics above and are not particularly high.
+Since imputation in multivariate analysis makes as many assumptions, I am not sure whether it would be beneficial compared to listwise deletion and its assumption of being missing at random.
 
@@
                   fml =  c(n_dk, n_ref, n_tot) ~ 
-                    duration + interference + showcards + notprimlang +
+                    n_applicable + interference + showcards + notprimlang +
                       gndrmatch + resolder + intolder +
@@
   filter(var != ".theta") %>%
-  filter(var %in% c("duration", "interference", "showcards", "notprimlang",
-                    "gndrmatch", "resolder", "intolder")) %>%
+  filter(var %in% c("duration", "n_applicable", "interference", "showcards", 
+                    "notprimlang", "gndrmatch", "resolder", "intolder")) %>%
   mutate(irr = exp(estimate),
@@
 Although I removed all NAs before, there are different numbers of observations for each dependent variable.
-This is due to some interviewers having no within-variance (that have only respondents with not a single item nonresponse).
+This is due to some interviewers having no within-variance (that have only respondents with no item nonresponse).
+Since a within-estimator needs within-variance, interviewers without within-variance are excluded from the analysis.
+This is particularly often the case for the number of refusals since the number of refusals is relatively low.
+Although such a large loss of observations is indeed unfortunate, they again can be assumed to be random and therefore not bias the results.
 
@@
 Wooldridge (1999) claims that fixed effects negative binomial regression (FENB) is not a true fixed effects estimator and that overdispersion is not of much concern in these models.
-He therefore advises to always use FE Poisson regression (FEPois) over FENB.
+He advises always using FE Poisson regression (FEPois) over FENB.
 
@@
 
-I compare estimates and diagnostic plots for both regression models using the total sum of item nonresponse as the dependent variable to assess reliability of my results.
+I compare estimates and diagnostic plots for both regression models using the total sum of item nonresponse as the dependent variable to assess the reliability of my results.
 
@@
                   fml =  n_tot ~
-                    duration + interference + showcards + notprimlang +
+                    n_applicable + interference + showcards + notprimlang +
                       gndrmatch + resolder + intolder +
@@
 While not perfect, the deviance residuals of the FENB model are much closer to normally distributed than FEPois.
-And the unconditional FENB by Allison and Waterman (2002) implemented in the fixest package is a true fixed effects model unlike previous versions.
+And the unconditional FENB by Allison and Waterman (2002) implemented in the fixest package is a true fixed effects model, unlike previous versions.
 It has an incidental parameter problem though, but it is usually very small.
@@
 Coefficients differ slightly between models but signs and significance are largely identical.
-I therefore settle on the FENB model in the paper.
-
-Clustered standard errors are the default in the fixest package.
-Normal parametric standard errors tend to be way too large (Cameron and Trivedi 2013).
-I replicated the regressions in Stata using bootstrap standard errors (not shown) and only get marginally different results, likely from differences in the optimisation algorithm.
+I, therefore, settle on the FENB model in the paper.
 
@@
 
-Allison, Paul D. and Richard P. Waterman (2002). “Fixed-Effects Negative Binomial Regression Models”. In: Sociological Methodology 32.1, pp. 247–265. doi: 10.1111/14679531.00117.
+Allison, Paul D. and Richard P. Waterman (2002). “Fixed-Effects Negative Binomial Regression Models”. In: Sociological Methodology 32.1, pp. 247–265. https://doi.org/10.1111/14679531.00117.
 
@@
 
-Wooldridge, Jeffrey M. (1999). “Distribution-Free Estimation of Some Nonlinear Panel Data Models”. In: Journal of Econometrics 90.1, pp. 77–97. doi: 10.1016/S0304-4076(98)00033-5.
+Wooldridge, Jeffrey M. (1999). “Distribution-Free Estimation of Some Nonlinear Panel Data Models”. In: Journal of Econometrics 90.1, pp. 77–97. https://doi.org/10.1016/S0304-4076(98)00033-5.

### 2024-01-31 inr_ESS9_text.tex
revised text after second round of reviews

@@
 \title{How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey}
-\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}
+%\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}
 
@@
 \begin{abstract}
-I review the literature on item nonresponse in surveys and based on this review, I extend the satisficing model with respondents' privacy concerns to incorporate all relevant aspects of the response process for item nonresponse.
+I review the literature on item nonresponse in surveys.
+Based on this review, I extend the satisficing model with respondents' privacy concerns to incorporate all relevant aspects of the response process for item nonresponse.
 I review proposed strategies to reduce item nonresponse and test selected strategies.
-Results suggest that boosting respondents' use of showcards, shorter interview durations and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse.
-Other people present during the interview only has a small association with the number of refusals.
-Matching age and gender of respondents and interviewers appears not to be a worthwile strategy.
+Results suggest that boosting respondents' use of showcards and interviewing in the respondents' primary language might be promising ways to reduce item nonresponse.
+Other people present during the interview have only a small association with the number of refusals.
+Matching the age and gender of respondents and interviewers appears not to be a worthwhile strategy.
 \end{abstract}
@@
 They decrease the effective sample size and can introduce bias to estimates if the causes for missingness are related to the item or respondent characteristics \citep{deleeuwPreventionTreatmentItem2003}.
-Although missing data are relatively rare in most single items, they can add up to a considerable loss of observations in multivariate analyses.
+Although missing data are rare in most single items, they can add up to a considerable loss of observations in multivariate analyses.
+Item nonresponse can also be seen as an indicator of overall data quality since it can result from \textit{satisficing} \citep{krosnickResponseStrategiesCoping1991}.
+Satisficing means that the respondent is giving a satisfactory answer instead of the best one.
 Due to these harmful effects of missing data, one objective of survey researchers is to keep their prevalence as low as possible.
@@
 
-I focus on the respondent as a source of missing data by answering with ``Don't know'' (DK) or refusing to answer.
-(\textit{item nonresponse}).
-Item nonresponse can also be seen as an indicator of overall data quality since it can be the result of \textit{satisficing} \citep{krosnickResponseStrategiesCoping1991}.
-Satisficing means that the respondent is giving a satisfactory answer instead of the best one possible.
+I review the existing literature on item nonresponse and extend the satisficing model based on this review to include privacy concerns resulting in an encompassing model of item nonresponse: the probability of item nonresponse depends on the task difficulty of the item(s) divided by the product of ability and motivation of the respondent or the respondents' privacy concerns.
+This means that higher difficulty results in more item nonresponse and higher ability and motivation in less item nonresponse.
+Difficulty, ability, and motivation are separate from privacy concerns, e.g.
+due to item sensitivity or general mistrust.
+Privacy concerns are relevant when deciding whether to disclose information and do not influence the cognitive burden of retrieving the answer.
 
-I review the existing literature on item nonresponse and extend the satisficing model based on this review to include privacy concerns resulting in an encompassing model of item nonresponse: the probability of item nonresponse depends on the task difficulty of the item(s) divided by the product of ability and motivation of the respondent or the respondents' privacy concerns.
-This means that higher difficulty results in more item nonresponse, higher ability and motivation in less item nonresponse.
-Difficulty, ability and motivation are separate from privacy concerns, e.g.
-due to item sensitivity or general mistrust.
-Privacy concerns are relevant when deciding whether to disclose an already formed answer and does not influence the cognitive burden of retrieving the answer.
+I then turn to practical strategies that could be used to decrease item nonresponse by reviewing proposed strategies.
+In the empirical part of the paper, I compare the effects of selected strategies.
+Promoting the use of showcards and translating questionnaires appear to be most promising.
+Both of those reduce the cognitive burden of respondents.
+Matching respondents' and interviewers' gender and age does not reduce item nonresponse.
+They could have influenced item nonresponse if respondents are more willing to share private information with interviewers similar to themselves.
+The results are not causal though.
+However, the results indicate that strategies aiming at a lower cognitive burden for respondents are our best guess to improve data quality.
 
-In the empirical part of the paper, I explore a few practical ideas to reduce item nonresponse.
-Promoting the use of showcards, shorter interviews and tranlating questionnaires might be promising strategies.
-All of those reduce the cognitive burden of respondents.
-The effects are not causal though because I am not able to sufficiently control for respondent ability.
-Matching respondents' and interviewers' gender and age seem not to be efficient tools to reduce item nonresponse.
-They could have influenced item nonresponse if respondents are more willing to share private information with interviewers that are similar to them.
-
-This introduction follows a short theoretical discussion presenting a model for the probability of item nonresponse based on satisficing.
+This introduction follows a short theoretical discussion presenting a theoretical model for the probability of item nonresponse based on satisficing.
 Structured by this model, I review the literature on strategies how to reduce item nonresponse and test a selection using the \textit{European Social Survey}.
@@
 
+Whether respondents answer a survey question and which answer they give is always a cognitive process taking place at the very moment of the interview.
 The \textit{survey response process} \citep{tourangeauPsychologySurveyResponse2000} involves multiple steps on behalf of the respondent.
-They need to comprehend the question, retrieve information from memory, eventually they need to judge these information, map them onto the response options and perhaps edit the response due to sensitivity or social desirability.
-Respondents can take these steps in order or jump back and forth, e.g.
+They need to comprehend the question, retrieve information from memory, eventually judge this information, map them onto the response options, and perhaps edit the response due to sensitivity or social desirability.
+Respondents will most likely take these steps in order but they can jump back and forth, e.g.
 if they need to form an opinion on the spot.
@@
 
-The two types of item nonresponse might be related to different steps of the survey response process.
-Refusals are more likely to stem from the editing step when respondents do not want to answer although they could.
-They might feel an item is too sensitive or do not trust the interviewer with this information.
-DK is more likely to come about when the respondent cannot answer in the first place.
+The two types of item nonresponse, "Don't know" and refusal, might be related to different steps of the survey response process though.
+Refusals are likely introduced in the editing step when respondents do not want to answer a question although they could.
+They may find certain information to be too sensitive or they may not feel comfortable sharing it with the interviewer due to a lack of trust.
+DK is likely as an answer when the respondent cannot give a substantive answer.
 Either the respondent cannot answer because they do not know about the content of the question or are unable to remember an event \citep{beattyAnswerNotAnwer2002, turnerCanResponseLatencies2015}.
-Or they do not perceive it worth the effort to come up with an answer and satisfice \citep{krosnickResponseStrategiesCoping1991}.
-But \citet{shoemakerItemNonresponseDistinguishing2002} have shown that higher mental effort is related to more refusals as well.
-And conversely, it is plausible that DK is used as a more polite way to refuse.
-I, therefore, do not distinguish between them in the following sections even though they often show different strengths of predictors \citep{silberEffectsQuestionRespondent2021}.
+In this situation, DK is a valid response and does not constitute a problem for data quality.
+On the other hand, they might not see value in putting in the effort to give an optimal response and \textit{satisfice} \citep{krosnickResponseStrategiesCoping1991}.
+Satisficing refers to various shortcuts (heuristics) that survey respondents can take when answering questions.
+One of these shortcuts is item nonresponse.
+The data collected in this case are of lower quality.
 
-Ability has been the explanation for the differences often found in item nonresponse by education, age, physical and mental health \citep{colsherDataQualityAge1989, pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}.
-Women and ethnic minority members are often found to have more item nonresponse \citep{kupekDeterminantsItemNonresponse1998, pickeryImpactRespondentInterviewer1998}.
-For the latter, on average lower literacy and lower command of the majority language are hypothesised to cause more item nonresponse \citep{kupekDeterminantsItemNonresponse1998}.
+However, \citet{shoemakerItemNonresponseDistinguishing2002}  have
+shown that higher mental effort is related to more refusals as well. And
+conversely, it is plausible that DK is used as a more polite way to refuse.
+In the following sections, I will therefore not distinguish between refusal and DK even though they may have varying strengths of predictors \citep{silberEffectsQuestionRespondent2021}.
+Similarly, I will not consider the unproblematic case of DK as a genuine answer although differentiating between the two meanings of DK might be relevant for substantive analyses.
+
+In the continuation of this section, I will discuss theoretical concepts that impact the likelihood of item nonresponse in the cognitive process of response formation.
+Later, I will combine these concepts into a theoretical model based on satisficing \citep{krosnickResponseStrategiesCoping1991}.
+
+When it comes to item nonresponse, the key concept is the \textit{ability} of the respondents to carry out cognitive tasks.
+Differences in cognitive abilities are the main explanation for differences in item nonresponse across education, age, and health \citep{colsherDataQualityAge1989, pickeryImpactRespondentInterviewer1998, deleeuwPreventionTreatmentItem2003, messerDeterminantsItemNonresponse2012, silberEffectsQuestionRespondent2021}.
+Ethnic minorities tend to have a higher rate of item nonresponse likely caused by lower literacy and worse command of the majority language \citep{kupekDeterminantsItemNonresponse1998, pickeryImpactRespondentInterviewer1998}.
 \citet{meitingerPowerCultureItem2020} conclude that item nonresponse reflects broader social inequalities in abilities and access to information.
-The ability hypothesis is also directly supported by correlations between item nonresponse and measures of intelligence \citep{hedengrenDogThatDidn2012}.
+The ability hypothesis is directly supported by correlations between item nonresponse and measures of intelligence \citep{hedengrenDogThatDidn2012}.
 
-More difficult or unclear questions typically have higher levels of item nonresponse \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.
-Since demographic questions can often be recalled very simply by respondents, they often show less item nonresponse than attitudinal and behavioural questions where the answers often do not exist before before getting asked the question \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.
+The second relevant concept is \textit{task difficulty}.
+When questions are more difficult or unclear, they tend to have higher rates of nonresponse \citep{holbrookImpactQuestionRespondent2006, messerDeterminantsItemNonresponse2012, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.
+Demographic questions are usually easier for respondents to remember, resulting in lower rates of item nonresponse compared to attitudinal and behavioral questions, which may require respondents to formulate an answer on the spot \citep{olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.
 
-Even when able to perform the task at hand, respondents not necessarily want to.
-They need to perceive the effort as worth it. Differences in \textit{motivation} are an explanation
-for less item nonresponse of respondents that are more interested in the topic of the survey \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}.
-Item nonresponse is related to measures of conscientiousness as well \citep{hedengrenDogThatDidn2012}.
+Even if people can complete a task, they may not want to do it unless they feel that the effort is worthwhile.
+They need to have the \textit{motivation} to provide an optimal response.
+That explains why people who are more interested in the topic of a survey are less likely to leave items unanswered \citep{kochItemNonresponseEuropean2009, silberEffectsQuestionRespondent2021}.
+Item nonresponse is linked to conscientiousness measures as well \citep{hedengrenDogThatDidn2012}.
 
-Motivation can also affect the editing process.
-For example, respondents (and interviewers) might deliberately not answer screening and filtering questions to reduce survey burden \citep{tourangeauMotivatedMisreportingShaping2015}.
-When editing an answer, concerns of privacy\footnote{I use this label to subsume general concerns about privacy towards the survey (e.g.
-about data processing, anonymity) and not wanting to answer single sensitive questions.} come into play.
-Am I going to be judged by the interviewer when answering honestly?
-Are my data really secure and confidential?
-This is a particular problem for questions perceived as more intrusive \citep{tourangeauSensitiveQuestionsSurveys2007}.
-Most prominently, this affects questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998}, which often show particularly high levels of item nonresponse.
-High item nonresponse is often used as an indicator for question sensitivity \citep{tourangeauSensitiveQuestionsSurveys2007}.
-Reluctancy and overall skepticism towards surveys and science, privacy concerns and generalised mistrust are traits that result in more nonresponse \citep{silberEffectsQuestionRespondent2021}.
+The editing process can also be influenced by motivation.
+For instance, respondents and interviewers may choose not to answer screening and filtering questions on purpose to lessen the survey workload \citep{tourangeauMotivatedMisreportingShaping2015}.
+This statement pertains only to data collections where the respondents know or can guess which questions serve as filters though.
+
+When editing an answer, respondents may have concerns about their \textit{privacy}\footnote{ I use this label to encompass overall privacy concerns related to the survey, such as data processing and anonymity, as well as the desire to avoid answering specific sensitive questions}.
+Will the interviewer judge me if I answer truthfully?
+Can I trust that my data will be kept secure and confidential?
+This is a particular problem for questions perceived as intrusive \citep{tourangeauSensitiveQuestionsSurveys2007} like questions on income \citep{yanTrendsIncomeNonresponse2010} and sexual behaviour \citep{kupekDeterminantsItemNonresponse1998}, which often show particularly high levels of item nonresponse.
+However, when it comes to attitude questions about controversial political issues such as immigration, there tends to be more item nonresponse as well \citep{piekutSurveyNonresponseAttitudes2021}.
+Item nonresponse is indeed frequently used as a measure of question sensitivity \citep{tourangeauSensitiveQuestionsSurveys2007}.
+Respondents will likely have such privacy concerns immediately when they hear a sensitive question and jump from the comprehension stage to the editing stage in the survey response process \citep{tourangeauPsychologySurveyResponse2000}.
+They probably refuse to answer before an honest answer has been formed.
+Increased item nonresponse is associated with reluctance and skepticism towards surveys and science, general privacy concerns, and mistrust \citep{silberEffectsQuestionRespondent2021}.
 
@@
 
-I have thus identified four constructs related to the probability of item nonresponse under which most research can be subsumed: cognitive ability, task difficulty, motivation and privacy concerns.
+Based on the reviewed literature, I have identified four concepts that affect the probability of item nonresponse in surveys: cognitive ability, task difficulty, motivation, and privacy concerns.
 However, as \citet{krosnickResponseStrategiesCoping1991} already hypothesised, these concepts are interrelated in their effect on item nonresponse.
@@
 The resulting fraction of difficulty by ability represents the relative mental effort to answer a question.
-To what point the relative mental effort is worth it depends on the respondents' motivation.
-A highly motivated respondent even answeres the difficult questions.
-The probability of nonresponse decreases with motivation.
-\citet[225]{krosnickResponseStrategiesCoping1991} has summarised these relations into a formula of the probability of satisficing.
-Satisficing refers to respondents lowering their standards in answering a question and only giving a satisfactory instead of an optimal answer to cut down on cognitive effort.
+And a highly motivated respondent answers even difficult questions.
+\citet[225]{krosnickResponseStrategiesCoping1991} formalised the probability of satisficing.
+
 Item nonresponse is such a satisficing strategy.
-There are situations where DKs are genuine though \citep{turnerCanResponseLatencies2015}.
-This is nonetheless a product of the respondents' ability and thereby included in the formula.
+Additionally, higher privacy concerns lead to more item nonresponse as well.
+Since this relates to another step in the survey response process, namely editing rather than comprehension, retrieval or judging, I postulate it to be independent from the other concepts.
 
-Additionally, higher privacy concerns lead to more item nonresponse.
-Since this relates to another step in the survey response process, namely editing rather than comprehension, retrieval or judging, I postulate it to be independent from the other concepts.
-I therefore suggest extending the formula by Krosnick as follows:
+For a complete theoretical model of harmful item nonresponse, privacy concerns therefore need to be added to the model by \citet{krosnickResponseStrategiesCoping1991}.
+As these concepts are (partially) interrelated and have a nonlinear relationship to item nonresponse, it is useful to formalize and summarize their relationship as follows:
 
-$$ P_{INR} \ = \ max \left( \frac{difficulty}{ability \ * \ motivation} \ , \ privacy \right) $$
+$$ P_{INR} \ = f\left(\ max \left( \frac{difficulty}{ability \ * \ motivation} \ , \ privacy \right)\right) $$
 
-\textit{The probability of a nonsubstantive answer by the respondent is given by the task difficulty
-divided by the product of ability and motivation or the respondents' privacy concerns, whichever
-is higher.}
+\textit{The probability of an ingenuine nonsubstantive answer on behalf of the respondent is a function of the task difficulty divided by the ability and motivation of the respondent or the respondents' privacy concerns, whichever is higher.}
 
-This formula specifically addresses item nonresponse for a single item but its meaning is adaptable to every level of a survey.
+Please note that this theoretical model is not able to and not intended to predict the probability of item nonresponse in a given item, as highlighted by the fact that it is an undefined function.
+Therefore, the individual concepts do not require measurement.
+The maximum function emphasizes that there are two independent mechanisms, and only the dominant one will impact item nonresponse at a time.
+This theoretical model specifically addresses item nonresponse for a single item but its meaning is adaptable to every level of a survey.
 
@@
 
-Based on these four concepts, one can theorise how to decrease item nonresponse.
-Some of the following possibilities may seem obvious and are established standards not only because of their potential relationship to item nonresponse but to ensure quality of substantial answers as well.
-Others might reduce item nonresponse but they could have negative consequences for other parts of \textit{total survey error}, the combined effect of all error sources in a survey \citep{grovesTotalSurveyError2010}.
+With these four concepts in mind, we can develop strategies to reduce item nonresponse and ensure better data quality in our surveys.
+Some of the following strategies may seem obvious and are already established standards in survey design not only because of their potential relationship to item nonresponse but to ensure the quality of substantial answers as well.
+Others might reduce item nonresponse but they could have negative consequences for other parts of total survey error, the combined effect of all error sources in a survey \citep{grovesTotalSurveyError2010}.
 They require a trade-off before implementation.
 
-I have structured this review of approaches to reduce item nonresponse by the respective concepts they target.
+I have structured this review of strategies to reduce item nonresponse by the respective concepts they target.
 
@@
 
-Task difficulty is predominantly influenced by design choices regarding single questions and their type of expected response.
-For the design of questions, much typical advice from question design aims to reduce task difficulty: asking short questions with simple structures, avoiding ambiguity and fuzzy concepts and not asking respondents to do multiple things at once.
-For a more comprehensive review on advice on question and questionnaire design, see for example \citet{smythDesigningQuestionsQuestionnaires2016}.
-But the type of question is also associated with task difficulty: open-ended questions, being allowed to select multiple options, more response categories and being asked to order categories are associated with higher item nonresponse \citep{schumanOpenClosedQuestion1979, holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.
-Additional measures to decrease difficulty can be visual helpers like images or showcards.
-With showcards, respondents do not need to remember all response categories while thinking about the question.
-But there is not much empirical work on the effect of showcards on item nonresponse.
-\citet{holbrookInterviewerErrorsHelp2016} found that questions using showcards had more item nonresponse although this is likely because they are only used with more difficult questions.
-In the context of the ESS, showcards do not seem to affect the distribution of substantive responses in experiments \citep{jackleAssessingEffectData2010} but item nonresponse was not investigated.
+The level of difficulty of a task is largely determined by how the questions are designed and what type of answer is expected.
+To make tasks easier, it is recommended to ask short, straightforward questions that avoid any confusion or unclear concepts.
+Asking respondents to complete multiple tasks at once should also be avoided.
+For a more thorough guide on how to design questions and questionnaires, see e.g.
+\citet{smythDesigningQuestionsQuestionnaires2016}.
+The difficulty of a task is related to the type of question as well.
+Questions that are open-ended or allow for multiple options and ordering of categories are more likely to result in higher nonresponse rates than closed single choice items \citep{schumanOpenClosedQuestion1979, holbrookImpactQuestionRespondent2006, holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019, silberEffectsQuestionRespondent2021}.
+To make it easier for respondents, visual aids like images or showcards can be used.
+Showcards eliminate the need for respondents to recall all response categories while answering a question.
+However, there is limited research on how showcards affect item nonresponse.
+According to a study by \citet{holbrookInterviewerErrorsHelp2016}, using showcards in survey questions led to more unanswered items.
+However, this may have been because showcards were only used for more challenging questions.
+In the European Social Survey (ESS), showcards do not appear to impact the distribution of meaningful responses in survey experiments, as noted by \citep{jackleAssessingEffectData2010}, although they did not investigate item nonresponse.
 
-The question level is very well understood and constitutes the largest part of the variance in item nonresponse.
-This highlights the importance of the single question for overall data quality \citep{olsonEffectsRespondentQuestion2019}.
+How question design affects levels of item nonresponse is very well understood and differences between questions constitute the largest part of the variance in item nonresponse \citep{olsonEffectsRespondentQuestion2019}.
+This highlights the importance of the single question for overall data quality.
 
-But not only single questions but the whole questionnaire needs to be designed as simply
-as possible.
-Questionnaires with changes in response categories, routing and filtering are associated with higher nonresponse \citep{messerDeterminantsItemNonresponse2012}.
-Grouping questions by topic could reduce mental effort and therefore item nonresponse but it also increases the likelihood of non-differentiation between the items \citep{krosnickResponseStrategiesCoping1991}.
-Offering DK and refusal explicitly increases their use because respondents are more aware of the “easy way out” \citep{schumanAssessmentNoOpinion1979, beattyAnswerNotAnwer2002}.
+It is important to design the entire questionnaire as simply as possible, not just the individual questions.
+Questionnaires that include changes in response scales, routing, and filtering tend to result in higher rates of nonresponse \citep{messerDeterminantsItemNonresponse2012}.
+However, routing and filtering should not increase difficulty in computer-assisted modes.
+Grouping questions by topic could reduce the required mental effort and item nonresponse but it also increases the likelihood of non-differentiation between items \citep{krosnickResponseStrategiesCoping1991}.
+Explicitly offering DK and refusal options can increase their use, as it makes respondents more aware of the possibility of an "easy way out" \citep{schumanAssessmentNoOpinion1979, beattyAnswerNotAnwer2002}.
 
-Task difficulty can be reduced for language minorities by providing the questionnaire in more languages.
-However, this is expensive and can reduce comparability of cases.
+To reduce task difficulty for members of language minorities, the questionnaire can be translated so that respondents can take the interview in the language they are most proficient in.
+But translating questionnaires can be costly and may affect the comparability of cases.
 For a review on comparability in cross-cultural surveys, see e.g.
@@
 
+To enhance the quality of survey design, identify any errors, and ensure that respondents can complete the required tasks, it is recommended to thoroughly review the questionnaire and its implementation for data collection.
+Common methods for doing so include conducting reviews and pilot studies.
+
 \subsection{Ability}
 
-To make sure that the respondent is able to answer the questions, interviews should be conducted in situations where understanding and concentration is as good as possible.
-Other people present during the interview, for example, could be a source of distraction.
-Empirical investigations have not been conclusive though \citep{kupekDeterminantsItemNonresponse1998, tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}.
-With longer interviews, respondents might get tired \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.
+While we cannot alter the general cognitive ability of our respondents, we can influence their ability to answer survey questions at the time of participation.
+To ensure a productive interview, it is important to choose an environment that encourages focus and clear communication.
+If possible, opt for quiet and not distracting locations at appropriate times.
+Having other people present during an interview can be distracting, but there is no conclusive evidence to support this claim \citep{kupekDeterminantsItemNonresponse1998, tuSocialDistanceRespondent2007, silberEffectsQuestionRespondent2021}.
+Respondents may become fatigued during lengthy interviews \citep{holbrookInterviewerErrorsHelp2016, olsonEffectsRespondentQuestion2019}.
 
@@
 
-Motivation could decrease over the course of the interview as well.
-Different page layouts and progress bars might help with that in web surveys but evidence is mixed \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}.
-The
-literature on cooperation enhancement, e.g.
-using incentives, has focussed on unit nonresponse.
-Some of these ideas might be applicable to increase respondent motivation within the interview as well.
+Motivation could decrease throughout the interview as well.
+While web surveys have used different page layouts and progress bars to combat this issue, the effectiveness of these methods is uncertain \citep{peytchevWebSurveyDesign2006, yanShouldStayShould2011, sarrafSurveyPageLength2014}.
+The research on cooperation enhancement, such as through incentives, has mainly focused on unit nonresponse.
+However, some of these methods could also be effective in increasing item nonresponse.
 After all, unit and item nonresponse are linked: respondents that initially refused to participate have higher levels of item nonresponse \citep{yanRelationUnitNonresponse2010, frickerExaminingRelationshipNonresponse2010}.
@@
 
-To tackle reluctancy to answer due to privacy concerns, survey researchers need to make sure that the respondents' data are secure and anonymous and communicate that.
-Explaining why the data are collected, how they are processed and what is done to protect their privacy is also ethically advisable and often legally required.
+Survey researchers should address privacy concerns to encourage respondents to answer by ensuring the security and anonymity of their data.
+It is important to communicate why the data is collected, how it will be processed, and how privacy is protected.
+This is not only ethically advisable but also often a legal requirement.
 
-In face-to-face surveys, establishing trust depends on the respondent-interviewer dynamics.
-Based on the hypothesis that respondents might perceive interviewers similar to themselves as more trustworthy, scholars have assessed the impact of matching their characteristics.
-\citet{vercruyssenEffectSociodemographicMis2017} find less item nonresponse when matching age
-of interviewers and respondents and matching gender reduces item nonresponse for males and increases it for females.
-Matching education has no effect \citep{vercruyssenEffectSociodemographicMis2017, silberEffectsQuestionRespondent2021}.
-\citet{tuSocialDistanceRespondent2007} on the other hand find age and education matching are potent predictors of item nonresponse.
+When conducting face-to-face surveys, the trust between the respondent and interviewer is influenced by their relationship.
+Scholars have hypothesized that respondents are more likely to trust interviewers who they perceive to be similar to themselves.
+To test this hypothesis, studies have been conducted to examine the impact of matching characteristics between the respondent and interviewer.
+\citet{vercruyssenEffectSociodemographicMis2017} observe less item nonresponse when interviewers and respondents are matched in age.
+Additionally, matching gender reduces item nonresponse for males but increases it for females.
+\citet{piekutSurveyNonresponseAttitudes2021} found female interviewers experienced higher rates of nonresponse but there was no significant correlation between the gender of the interviewer and the gender of the respondent.
+\citet{silberEffectsQuestionRespondent2021} found that education matching has no effect while \citet{tuSocialDistanceRespondent2007} find age and education matching to be potent predictors of item nonresponse.
+A test that could be interesting to conduct is whether pairing interviewers and respondents who share the same immigration status and/or ethnicity would make a difference.
+Immigrants tend to have higher levels of item nonresponse, language barriers, and I could imagine that some of them may be mistrustful towards interviewers due to racist experiences and a fear of discrimination.
 
@@
 
-In this review, I have stressed the role of the respondent and how design choices can increase the probability of a substantial answer.
-But in face-to-face surveys, the interviewer can be a source of item nonresponse as well.
-This is left for further research.
-I proceed with testing a few proposed strategies for item nonresponse reduction based on this review.
+So far, this article has reviewed and summarized the literature on item nonresponse in surveys.
+I have suggested a theoretical model, an extension of the satisficing model by \citet{krosnickResponseStrategiesCoping1991}, as a conceptual summary that can inform our survey design and I have reviewed strategies to reduce item nonresponse and categorized them accordingly.
+In the remainder of the article, I am going to test a few selected strategies to reduce item nonresponse derived from the theoretical model and the literature review.
+All of these strategies could change at least one of the four concepts from the theoretical model and therefore could have a causal connection to item nonresponse.
+Whether these strategies actually do change the associated concepts and how strongly their effect translates into changes in item nonresponse will be central to my empirical analysis.
 
-Over time, respondents might lose the ability to concentrate and motivation.
-I therefore expect more item nonresponse, the longer the interview takes.
-If this is true, making questionnaires only as long as necessary is a strategy to reduce item nonresponse and maintain high data quality.
-However, the duration of the interview might also be dependent on the respondents' ability.
-Overall ability of the respondent needs to be controlled for in the model.
+\begin{enumerate}
+ \item During an interview, respondents may experience a decrease in concentration and motivation to answer questions as time goes on.
+As a result, item nonresponse may become more common the longer the interview lasts.
+To maintain high data quality and reduce item nonresponse, it might be advisable to keep questionnaires as short as possible.
 
-Interferences and other people being present during the interview are expected to increase item nonresponse because they can be a source of distraction and respondents might not want to answer some questions in front of people they know.
-If true, interviews should be conducted without others present if possible.
+ \item It is likely that interferences and the presence of others during an interview can cause item nonresponse, as they may distract the respondent and make them hesitant to answer certain questions in front of people they know.
+As a result, it might be beneficial that interviews are conducted without other people present, if feasible.
 
-Showcards are intended to make the process of answering less burdensome because respondents do not need to remember the response options.
-If this has an effect on item nonresponse, showcards should be provided and their use promoted to ensure a high data quality.
-In this operationalisation, the interviewer assesses afterwards to what extent the respondent used the showcards.
-Respondent ability might influence how well they understand the procedures, ability needs to be controlled for again.
-The interviewer assessment is likely influenced by the overall impression the interviewer has from the interview and is therefore endogenous.
-The overall impression of the interviewer needs to be controlled for as well.
+ \item To make answering easier for respondents, showcards could be provided so they do not have to remember the response scale.
+Showcards would then lower the required cognitive effort and reduce item nonresponse.
 
-Language difficulties are likely to inhibit the response process.
-For respondents that primarily speak a different language at home than the interview is conducted in, the task might be more difficult.
-To ensure data quality from these respondents, translating the questionnaire is an option (although costly and comes with problems of comparability).
-Respondents for whom this is true are likely migrant groups that have lower levels of education than the native population.
-Therefore, ability needs to be controlled for.
+ \item Respondents who primarily speak a different language at home may face difficulties in the response process.
+To ensure data quality from these respondents, one option is to translate the questionnaire, although this can be costly and may present comparability problems.
 
-Previous research suggests that respondents might be more willing to answer to socially similar interviewers.
-I test this proposition for gender and age.
-If true and the sampling frame contains some demographic information, survey agencies could assign interviewers based on these demographics.
-Whether the gender of respondent and interviewer matches can be seen as random while the probability of the respondent being older or younger than the interviewer is dependent on the respondents' age, which is correlated with ability.
-Therefore, respondent age needs to be controlled for \footnote{Interviewer age is accounted for in the fixed effects.}.
+ \item According to previous studies, people may feel more comfortable answering questions from interviewers who share similar social characteristics, such as gender and age.
+If true, survey agencies could assign interviewers based on demographic information if it is available in the sampling frame.
+\end{enumerate}
 
-Table \ref{table_strat} summarises the strategies to be tested.
-Necessary controls are respondents' ability and specifically age and the interviewers' overall assessment of the interview.
-Based on the theory outlined above, I do not see that inclusion of any of these control variables or the other strategies are likely to distort the effect of another variable of interest.
-Therefore, I am going to test all effects in a single model.
+Table 1 summarises the selected strategies that I am going to test in my empirical analysis.
+The second column shows which concepts play a role in the hypothesized mechanism linking the respective strategy to item nonresponse.
+The third column gives the expected direction of the relationship between strategy and item nonresponse, e.g.
+the longer the questionnaire, the more item nonresponse.
+These are also the expected signs of the coefficients if the strategies work as imagined.
 
@@
 \hline
-Duration of the Interview & Ability, Motivation & positive \\
+Length of the Questionnaire & Ability, Motivation & positive \\
 Interference of the Interview & Ability, Privacy & positive \\
@@
 \end{tabular}
-\label{table_strat}
 \end{table}
@@
 
-Following this review and synthesis, I want to test the effectiveness of some potential strategies to reduce item nonresponse.
-I use the \textit{European Social Survey (ESS)} Round 9 collected in 2018 \citep{essericEuropeanSocialSurvey2019}.
-The ESS is a biannual face-to-face survey on attitudes and beliefs towards social and political topics in Europe.
+To test the effectiveness of some potential strategies to reduce item nonresponse, I use the \textit{European Social Survey (ESS)} Round 9 collected between August 2018 and January 2020 \citep{essericEuropeanSocialSurvey2019}.
+The ESS is a biannual face-to-face trend survey on attitudes and beliefs towards social and political topics in Europe established in 2001.
+In each country and round, the ESS draws a new random sample of the residential population of 15 years and older aiming for a minimum response rate of 70\%.
+Most countries use computer-assisted personal interviews for data collection and the questionnaire is designed to take about one hour.
 The data release 3.1 includes data from 49,519 respondents from 29 countries.
+For more information on the data, see the supplementary material or visit \href{https://www.europeansocialsurvey.org}{europeansocialsurvey.org}.
 
@@
 
-The three dependent variables are the sum of DK, the sum of refusals and the total sum of item nonresponse for every respondent.
-Item nonresponse is counted only in variables that are presented to all respondents, i.e.
-that are not affected by filtering questions.
-DK and refusal are not presented as response options in the ESS but are explicitly recorded as such.
-Interviewers are instructed to accept them without probe.
+The three dependent variables are the sum of DK, the sum of refusals, and the total sum of item nonresponse for every respondent.
+Non-responses are only counted for variables that are presented to all respondents and not affected by filtering questions.
+Respondents are not given the option to respond with DK or refusal, but interviewers are instructed to record them explicitly and without further probing.
+It is up to the interviewer to interpret a non-response as either a refusal or DK.
 
-\subsection{Independent Variables}
+Although my argument focuses on item nonresponse which is problematic for data quality opposed to DK as a genuine answer, I have not separated the two meanings in the analysis for two reasons.
+Firstly, distinguishing between these two meanings is often very challenging, and it requires a deep understanding of the specific question, which is not feasible for this general analysis.
+Secondly, an additional mechanism that generates item nonresponse may increase the overall variation in the dependent variables but if it is uncorrelated to the other mechanisms, no bias in estimates is to be expected.
+I do not know how the possibility of genuine DK answers could interfere with the other mechanisms.
+I therefore assume that they are uncorrelated.
 
-The independent variables of interest are discussed in the previous section.
+\subsection{Control Variables}
 
-To control for ability and the interviewers' assessment of the interview, I include education (operationalised by the ISCED scale), age and squared age of the respondent as well as the interviewers' assessment of how well the respondent understood the questions, to what extent the respondent answered to the best of their ability and how often they asked for clarifications.
+As my empirical analysis is concerned with strategies that potentially could be used in survey design and implementation to reduce problematic item nonresponse and therefore to increase data quality, it aims at \textit{causal inference} \citep{angristMostlyHarmlessEconometrics2009}: Do we expect a difference in item nonresponse if a strategy was implemented compared to the counterfactual when it was not implemented?
+Or in other words, does the implementation of a strategy \textit{cause} a net decrease in item nonresponse on average?
 
-More information on all variables including summary statistics can be found in the \href{https://doi.org/10.17605/OSF.IO/M83GY
-}{supplemental material}.
+To identify the average treatment effects of the selected strategies with cross-sectional survey data, I need to control for potential sources of bias in the effects of the strategies, other unrelated influences can be omitted.
+Such selection of controls always requires a sufficiently complete theory.
+In this case, the selection of controls can be based on the theoretical model outlined earlier.
 
-\subsection{Model}
+In my analysis, I need to control for respondents' ability as it is likely related to respondents' understanding of survey procedures like showcard use.
+Ability also needs to be controlled to estimate the effect of language differences as immigrant and minority groups in Europe typically differ in education compared to the majority groups.
+Ability is also related to the respondents' age and could therefore bias the effect of matching interviewers' and respondents' characteristics.
 
-I analyse the data using a negative binomial regression with interviewer fixed effects \citep{allisonFixedEffectsNegativeBinomial2002}\footnote{Count data models with fixed effects are quite debated \citep{wooldridgeDistributionfreeEstimationNonlinear1999, allisonFixedEffectsNegativeBinomial2002, cameronRegressionAnalysisCount2013}.
-For a justification of model choice, see the supplementary material.}.
-Standard errors are clustered by interviewer.
-Since the population of interest in this case are interviews and not societies, no weighting is applied.
+As the use of showcards is evaluated by the interviewer after the interview, the test of the effectiveness of showcards has an endogeneity problem.
+The overall impression the interviewer has of the respondent might influence the perception of showcard use.
+I, therefore, control for the interviewer's general impression of the interview.
+
+I am not aware of any mechanisms that could lead to biased estimates for the effects of interferences and other people being present during the interview as well as whether respondent and interviewer have matching gender.
+In summary, necessary controls are therefore respondents' ability and specifically age and the interviewers' overall assessment of the interview.
+Based on the theory outlined above, I do not expect that the inclusion of any of these control variables or the other strategies is likely to distort the effect of another variable of interest.
+Therefore and to be able to compare relative effect size, I am going to test all effects in a single regression model.
+
+However, identifying causal effects in regression modeling requires the \textit{conditional independence assumption} \citep[52ff]{angristMostlyHarmlessEconometrics2009} that all sources of bias are sufficiently controlled for.
+This is a strong assumption as it requires not only a sufficient theory (and the sufficiency of a theory is improvable) but also the operationalization, measurement, and functional form of the statistical model needs to be correct.
+This is never the case in social research \citep{martinThinkingStatistics2018}.
+Even though I have carefully selected the controls based on the presented theoretical model, I can only use proxies for the concepts I need to control for.
+I will therefore not speak of causal effects but of (conditional) associations as the point estimates can still be slightly off.
+Nonetheless, the regression estimates should reveal which strategies work and which are the most promising for implementation.
+Future experimental research could investigate the most promising strategies more thoroughly.
+
+\subsection{Independent Variables of Interest}
+
+I calculate the number of questions the respondent was asked by subtracting the number of items coded as not applicable from the total number of questions.
+Whether the interview was conducted in the respondent's primary language is a dummy variable generated from the metadata in which language the interview was conducted and the respondent's answer to the question of which language they primarily speak at home.
+Matching social characteristics are also dummy variables and generated from demographic information from the main questionnaire and the interviewer questionnaire.
+The interviewer questionnaire is a short questionnaire the interviewer fills out after completing the interview.
+For matching ages, I constructed two dummies whether the respondent is more than ten years older or younger.
+The reference category is whether the age difference is ten years maximum.
+I went for a cutoff difference of ten years to have a meaningful and visually perceivable difference in age and enough observations in all categories.
+The interviewer questionnaire also asks whether other people were present during the interview or not.
+And interviewers rate the respondents' use of showcards on a three-point scale: respondent used all the applicable showcards, respondent used only some applicable showcards, respondent refused/ was unable to use the showcards at all.
+I treat this latter variable as metric with higher values indicating more frequent use of showcards.
+
+To control for ability, I use education (operationalized by the ISCED scale), age, and squared age of the respondent as proxies.
+To dampen the endogeneity problem of interviewers' assessment of showcard use, I include the interviewers' assessment of how well the respondent understood the questions, to what extent the respondent answered to the best of their ability, and how often they asked for clarifications.
+They serve as proxies for ability as well.
+
+\subsection{Statistical Model}
+
+Since the dependent variables are count data and show the typically skewed distribution of count data, I analyze the data using a negative binomial regression with interviewer fixed effects \citep{allisonFixedEffectsNegativeBinomial2002}.
+The interviewer fixed effects are used to control for mean differences in interviewer behavior regarding accepting and recording item nonresponse.
+At the same time, they absorb variation between countries.
+Standard errors are clustered by the interviewer following recommended practice to prevent heteroscedasticity \citep[358f]{cameronRegressionAnalysisCount2013}.
+Since the population of interest are interviews, no weighting is applied.
 Missing data are deleted listwise.
 
-The analyses are carried out in \textit{R} \citep{rcoreteamLanguageEnvironmentStatistical2021} using the \textit{Tidyverse} \citep{wickhamWelcomeTidyverse2019} for data handling and graphics and the \textit{fixest} package \citep{bergeEfficientEstimationMaximum2018} to estimate the regressions.
-The code is available  in the \href{https://doi.org/10.17605/OSF.IO/M83GY
-}{supplemental material} as well.
+The analyses are carried out in \textit{R} \citep{rcoreteamLanguageEnvironmentStatistical2023} using the \textit{Tidyverse} \citep{wickhamWelcomeTidyverse2019} for data handling and graphics and the \textit{fixest} package \citep{bergeEfficientEstimationMaximum2018} to estimate the regressions.
+
+More information on the data, variables, summary statistics, all of the code used for preparation and analysis, and discussions on missing values and model specification are available in the  %\href{https://doi.org/10.17605/OSF.IO/M83GY}
+{supplemental material} [OSF LINK HIDDEN FOR REVIEW].
+
+Count data models with fixed effects are quite debated \citep{wooldridgeDistributionfreeEstimationNonlinear1999, allisonFixedEffectsNegativeBinomial2002, cameronRegressionAnalysisCount2013}.
+For a thorough discussion of model choice, see the supplementary material as well.
 
@@
 
-Figure \ref{fig_model} shows the incidence rate ratios (exponentiated coefficients) of the tested strategies.
-The exact coefficients of the tested strategies and of the control variables as well as standard errors and coefficients of model fit can be found in the table in the appendix.
+Figure 1 shows the incidence rate ratios (exponentiated coefficients) of the coefficients of interest.
+The coefficients to evaluate the tested strategies and of the control variables as well as standard errors and coefficients of model fit can be found in Table 2.
 
@@
 \centering
-\caption{Coefficients of the tested Strategies}
-\includegraphics[width=\linewidth]{results_inr_ess9_figure.pdf}
-\label{fig_model}
+\caption{Coefficients of Interest}
+\includegraphics[width=\linewidth]{results_inr_ess9.pdf}
 \end{figure}
 
-The most promising ways to reduce item nonresponse seem to be boosting the use of showcards, shorter interviews and tranlating questionnaires.
+{\small
+\include{results_inr_ess9}
+}
 
-With more showcard use as indicated by the interviewer (respondent used no showcards vs.
-used some showcards vs.
-used all showcards), the amount of DK reduces by about a quarter and the amount of refusal by about 18\% on average.
-With every additional thirty minutes interview duration, the number of DKs rises by about 21\% and the number of refusals rises by about 8\% on average.
-And compared to interviews conducted in the language the respondent primarily speaks at home, interviews conducted in a language different from the respondents' primary language at home show on average 42\% more DKs and 26\% more refusals.
+The most promising ways to reduce item nonresponse seem to be boosting the use of showcards and translating questionnaires.
+With more frequent showcard use as indicated by the interviewer, the amount of DK reduces by about a quarter and the amount of refusal by about 18\% on average.
+And compared to interviews conducted in the language the respondent primarily speaks at home, interviews conducted in a language different from the respondents' primary language show on average 42\% more DKs and 26\% more refusals.
 
@@
 This is not surprising since there are many more DKs than refusals.
-The effects on the number of refusals is typically weaker than the effect on DK but still present.
+The effects on the number of refusals are typically weaker than the effect on DK but still present.
 This supports the idea that refusals and DKs are not perfectly separate in their meaning but not identical as well.
@@
 Matching respondents' gender has a small positive effect on the number of DKs.
-This is contrary to expectations, which suggested that matching socio-demographics of interviewers and respondents leads to a more trusting interview situation and reduces item nonresponse.
+This is contrary to expectations, which suggested that matching the socio-demographics of interviewers and respondents leads to a more trusting interview situation and reduces item nonresponse.
 The effect on refusals is not significant but should be pronounced since this strategy is partly based on the privacy mechanism.
 Matching by age has no significant effect.
-Matching respondents and interviewers therefore seems not to be a promising strategy to reduce item nonresponse.
+Matching respondents and interviewers seems not to be a promising strategy to reduce item nonresponse.
 
-Other people present during the interview raises the amount of refusals by 14\% in line with the reasoning that respondents do not want to answer some questions in presence of others they know.
+Other people present during the interview raised the number of refusals by 14\% in line with the reasoning that respondents do not want to answer some questions in the presence of others they know.
 The effect on DK is not significant.
-Other people present might therefore influence item nonresponse more via privacy than distraction.
+Other people present might therefore influence item nonresponse more via privacy than a distraction.
 However, due to the relatively small number of refusals, the effect on the total item nonresponse is not significant.
+
+Contrary to expectation, the number of applicable items has a significantly negative effect on DK (and total item nonresponse).
+A respondent that has been asked 10 questions more has a 12\% lower average number of DK answers.
 
@@
 I have reviewed the literature on item nonresponse and extended the cognitive \textit{satisficing} model \citep{krosnickResponseStrategiesCoping1991} with concerns about privacy to encompass all aspects that can interfere with the response process in survey interviews.
-Based on this new model, I have reviewed possibilities to reduce item nonresponse in face-to-face surveys.
+Organizing our knowledge into such theoretical models highlights the interrelations between theoretical constructs which is necessary to reduce total error and is not achievable with piece-meal empirical studies.
+Based on this new model, I have reviewed possibilities to reduce item nonresponse, particularly in face-to-face surveys.
 
-In an empirical analysis using data from the \textit{European Social Survey Round 9}, I found that boosting the respondents' use of showcards, shortening interview durations and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse in face-to-face surveys.
+In an empirical analysis using data from the \textit{European Social Survey Round 9}, I found that boosting the respondents' use of showcards and conducting the interview in the respondents' primary language might be promising ways to reduce item nonresponse in face-to-face surveys.
 These strategies reduce the cognitive effort on behalf of respondents.
-Other people present during the interview is moderately associated with more refusals as well.
-Respondents are probably not willing to disclose private information in front of people they know.
-However, my hypothesis that respondents might trust interviewers more and share more private information if interviewer and respondent are socially similar has received no support: matching socio-demographic characteristics of interviewers and respondents seems not to be a worthwile strategy.
+Other people present during the interview are moderately associated with more refusals.
+Respondents are probably unwilling to disclose private information in front of people they know.
+However, my hypothesis that respondents might trust interviewers more and share more information if the interviewer and respondent are socially similar has received no support: matching the socio-demographic characteristics of interviewers and respondents seems not a worthwhile strategy.
+And surprisingly, longer questionnaires were associated with less item nonresponse.
+However, this might be related to a problem of operationalization.
+Most questions that might not apply to respondents are demographics asked at the end.
+But this would explain no association, but I observe a negative effect for which I do not have an explanation.
 
-I want to stress though that I do not claim causal effects.
 Although I carefully selected control variables, I cannot rule out violations of the \textit{conditional independence assumption} which is necessary to identify causal effects with regression analysis.
@@
 Respondents' ability is notoriously hard to measure in surveys and proxies like education and age that I have used as controls are not perfect.
+A second threat to the results is the endogeneity of some variables of interest, in particular, showcard use and others present during the interviews.
+They are measured in the interviewer questionnaire after the interview and are likely biased by the interviewers' overall assessment of the interview, including the amount of item nonresponse.
+I tried to control for that using other variables from the interviewer questionnaire.
+A third limitation of this analysis concerns the external generalizability of the results.
+We know that specific types of questions are more prone to item nonresponse, for example, opinions and sensitive questions.
+The results obtained here reflect the effects on item nonresponse especially on matters of opinion as this is the primary object of study of the ESS.
+While opinion surveys constitute a large share of surveys and the results should be generalizable to them, other types of surveys might have slightly different challenges concerning item nonresponse.
 
-A second threat to the results is the endogeneity of some variables of interest, in particular showcard use and others present during the interviews.
-They are measured in the interviewer questionnaire after the interview and are likely biased by the interviewers' overall assessment of the interview, including the amount of item nonresponse.
-I tried to control for that using other variables from the interviewer questionnaire but again, I cannot verify the conditional independence assumption to be true.
+I nonetheless see value in this analysis for two reasons.
+First, item nonresponse is relatively rare in single items and therefore difficult to study using survey experiments.
+Second, and more importantly, the empirical analysis aims to compare multiple potential strategies in their strength of effect (which is not possible using experiments).
+While it is difficult to assess the true causal effect of the strategies that do make a difference, the strategies that have no effects even in this biased analysis will likely not be successful.
+This analysis can provide a meaningful starting place for more rigorous tests of the most promising strategies and nonetheless inform survey design choices.
 
-In summary, this review and analysis can only serve as a \textit{current best guess} in how to think about item nonresponse and its reduction.
-Future research might be able to overcome some of the shortcomings of this and other previous studies.
+Although I analyze data from a face-to-face survey, I think it is important to anticipate some of the results and especially the implications of the theoretical model for the current shift to self-administrating modes of data collection.
+The results of my analysis highlight that the most promising strategies to decrease item nonresponse are tools that decrease task difficulty, like showcards and translating questionnaires.
+In self-administered modes, designing easy-to-use and clear questionnaires and page layouts will be important for item nonresponse.
+For paper-based modes, this will limit the options for routing and filtering.
+Specifically, offering refusal and DK as response options is an important design choice.
+Ethically, respondents need to have the option not to respond.
+On one hand, this will likely increase item nonresponse.
+On the other hand, forcing an answer will generate low-quality responses.
+In self-administered modes, we have less control over the interview situation, for example, whether other people are present.
+My analysis has shown that the latter is associated with more refusals.
+The absence of an interviewer reduces social desirability.
+No social desirability is often considered an advantage as respondents do not need to disclose information to a stranger.
+But no interviewer could also be a disadvantage as there might be a lower hurdle to satisfice.
+But the strategies based on the idea that respondents trust socially similar interviewers indicated that the presence of the interviewer might be less important for general levels of item nonresponse.
+Finally, self-administered modes can be conducted in multiple languages easily because we do not need interviewers that speak a minority language.
 
@@
 
-I want to thank Achim Koch, Johanna Mehltretter and especially Tobias Gummer for helpful comments in multiple stages of this project.
+[HIDDEN FOR REVIEW]
+
+%I want to thank Achim Koch, Johanna Mehltretter and especially Tobias Gummer for helpful comments in multiple stages of this project.
 I want to thank Marcel Kappes and Thomas Gautschi for methodological discussions.
 
-This work was supported by the University of Mannheim’s Graduate School of Economic and Social Sciences.
+%This work was supported by the University of Mannheim’s Graduate School of Economic and Social Sciences.
 
@@
 
-\newpage
-
-\include{results_inr_ess9_table}
-
 \end{document}

### 2024-02-06 inr_ESS9_text.tex
de-anonymised version after reviews

@@
 \title{How to reduce Item Nonresponse in Face-to-Face Surveys? A Review and Evidence from the European Social Survey}
-%\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}
+\author{Malte Grönemann \\ University of Mannheim, Mannheim, Germany}
 
@@
 
-More information on the data, variables, summary statistics, all of the code used for preparation and analysis, and discussions on missing values and model specification are available in the  %\href{https://doi.org/10.17605/OSF.IO/M83GY}
-{supplemental material} [OSF LINK HIDDEN FOR REVIEW].
+More information on the data, variables, summary statistics, all of the code used for preparation and analysis, and discussions on missing values and model specification are available in the  \href{https://doi.org/10.17605/OSF.IO/M83GY}{supplemental material}.
 
@@
 
-[HIDDEN FOR REVIEW]
-
-%I want to thank Achim Koch, Johanna Mehltretter and especially Tobias Gummer for helpful comments in multiple stages of this project.
+I want to thank Achim Koch, Johanna Mehltretter and especially Tobias Gummer for helpful comments in multiple stages of this project.
 I want to thank Marcel Kappes and Thomas Gautschi for methodological discussions.
 
-%This work was supported by the University of Mannheim’s Graduate School of Economic and Social Sciences.
+This work was supported by the University of Mannheim’s Graduate School of Economic and Social Sciences.
 
